\documentclass[12pt]{article}
\usepackage[]{accessibility}
\usepackage{hyperref,geometry,setspace}
%\usepackage[symbol*,perpage]{footmisc}
\geometry{width=6in,height=9in,centering}
% changes font to TeX Gyre Schola (Century Schoolbook)
\usepackage{tgschola}\usepackage[T1]{fontenc}



\newcounter{wjparagraph}[section]
\newcommand*{\wjparagraph}{\stepcounter{wjparagraph}~\marginpar{\tiny\arabic{wjparagraph} }}


\begin{document}
%\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
%\setcounter{footnote}{0}
\hypersetup{pdfinfo={Title={Pragmatism}, Author={William James}}, pdfborder = {0 0 0 0}}

\paragraph{About this text:} This includes the preface and lectures 1--3. It was prepared by P.D. Magnus from \href{http://www.gutenberg.org/ebooks/5116}{the Project Gutenberg edition}. Additions by P.D.M. are in bracketed footnotes. In order to facilitate comparison with other editions, paragraphs are numbered in the margin.

\begin{center}
\section*{Pragmatism:\\ A New Name for Some Old Ways of Thinking}

{\textsc{William James}}
\end{center}


\begin{center}
To the Memory of John Stuart Mill

from whom I first learned the pragmatic openness of mind and whom my
fancy likes to picture as our leader were he alive to-day.
\end{center}



\section*{Preface}

The lectures that follow were delivered at the Lowell Institute in
Boston in November and December, 1906, and in January, 1907, at
Columbia University, in New York. They are printed as delivered, without developments or notes. The pragmatic movement, so-called--- I do not like the name, but apparently it is too late to change it--- seems to have rather suddenly precipitated itself out of the air. A number of tendencies that have always existed in philosophy have all at once become conscious of themselves collectively, and of their combined mission; and this has occurred in so many countries, and from so many different points of view, that much unconcerted statement has resulted. I have sought to unify the picture as it presents itself to my own eyes, dealing in broad strokes, and avoiding minute controversy. Much futile controversy might have been avoided, I believe, if our critics had been willing to wait until we got our message fairly out.
\ldots\footnote{[Omitted here are ``a few references'' to works by John Dewey, F.C.S. Schiller, and others.]}

\medskip
%If my lectures interest any reader in the general subject, he will doubtless wish to read farther. I therefore give him a few references.

%In America, John Dewey's `Studies in Logical Theory' are the foundation. Read also by Dewey the articles in the Philosophical Review, vol. xv, pp. 113 and 465, in Mind, vol. xv, p. 293, and in the \emph{Journal of Philosophy}, vol. iv, p. 197.

%Probably the best statements to begin with however, are F. C. S. Schiller's in his `Studies in Humanism,' especially the essays numbered i, v, vi, vii, xviii and xix. His previous essays and in general the polemic literature of the subject are fully referred to in his footnotes.

%Furthermore, see G. Milhaud: le Rationnel, 1898, and the fine articles by Le Roy in the \emph{Revue de Metaphysique}, vols. 7, 8 and 9. Also articles by Blondel and de Sailly in the \emph{Annales de Philosophie Chretienne}, 4me Serie, vols. 2 and 3. Papini announces a book on \emph{Pragmatism}, in the French language, to be published very soon.

To avoid one misunderstanding at least, let me say that there is no logical connexion between pragmatism, as I understand it, and a doctrine which I have recently set forth as `radical empiricism.' The latter stands on its own feet. One may entirely reject it and still be a pragmatist.

~\hfill Harvard University, April, 1907.


\section*{Lecture I: The Present Dilemma in Philosophy}
\setcounter{wjparagraph}{0}

\wjparagraph
In the preface to that admirable collection of essays of his called \emph{Heretics}, Mr. Chesterton writes these words: ``There are some people--- and I am one of them--- who think that the most practical and important thing about a man is still his view of the universe. We think that for a landlady considering a lodger, it is important to know his income, but still more important to know his philosophy. We think that for a general about to fight an enemy, it is important to know the enemy's numbers, but still more important to know the enemy's philosophy. We think the question is not whether the theory of the cosmos affects matters, but whether, in the long run, anything else affects them.''

\wjparagraph
I think with Mr. Chesterton in this matter. I know that you, ladies and gentlemen, have a philosophy, each and all of you, and that the most interesting and important thing about you is the way in which it determines the perspective in your several worlds. You know the same of me. And yet I confess to a certain tremor at the audacity of the enterprise which I am about to begin. For the philosophy which is so important in each of us is not a technical matter; it is our more or less dumb sense of what life honestly and deeply means. It is only partly got from books; it is our individual way of just seeing and feeling the total push and pressure of the cosmos. I have no right to assume that many of you are students of the cosmos in the class-room sense, yet here I stand desirous of interesting you in a philosophy which to no small extent has to be technically treated. I wish to fill you with sympathy with a contemporaneous tendency in which I profoundly believe, and yet I have to talk like a professor to you who are not students. Whatever universe a professor believes in must at any rate be a universe that lends itself to lengthy discourse. A universe definable in two sentences is something for which the professorial intellect has no use. No faith in anything of that cheap kind! I have heard friends and colleagues try to popularize philosophy in this very hall, but they soon grew dry, and then technical, and the results were only partially encouraging. So my enterprise is a bold one. The founder of pragmatism himself recently gave a course of lectures at the Lowell Institute with that very word in its title--- flashes of brilliant light relieved against Cimmerian darkness!\footnote{[James is referring to lectures by Charles Sanders Peirce at Harvard in 1903.]} None of us, I fancy, understood \emph{all} that he said--- yet here I stand, making a very similar venture.

\wjparagraph
I risk it because the very lectures I speak of \emph{drew}--- they brought good audiences. There is, it must be confessed, a curious fascination in hearing deep things talked about, even tho neither we nor the disputants understand them. We get the problematic thrill, we feel the presence of the vastness. Let a controversy begin in a smoking-room anywhere, about free-will or God's omniscience, or good and evil, and see how everyone in the place pricks up his ears. Philosophy's results concern us all most vitally, and philosophy's queerest arguments tickle agreeably our sense of subtlety and ingenuity.

\wjparagraph
Believing in philosophy myself devoutly, and believing also that a kind of new dawn is breaking upon us philosophers, I feel impelled, \emph{per fas aut nefas},\footnote{[Latin for `by good or ill'; that is, `by any means'.]} to try to impart to you some news of the situation.

\wjparagraph
Philosophy is at once the most sublime and the most trivial of human pursuits. It works in the minutest crannies and it opens out the widest vistas. It `bakes no bread,' as has been said, but it can inspire our souls with courage; and repugnant as its manners, its doubting and challenging, its quibbling and dialectics, often are to common people, no one of us can get along without the far-flashing beams of light it sends over the world's perspectives. These illuminations at least, and the contrast-effects of darkness and mystery that accompany them, give to what it says an interest that is much more than professional.

\wjparagraph
The history of philosophy is to a great extent that of a certain clash of human temperaments. Undignified as such a treatment may seem to some of my colleagues, I shall have to take account of this clash and explain a good many of the divergencies of philosophers by it. Of whatever temperament a professional philosopher is, he tries when philosophizing to sink the fact of his temperament. Temperament is no conventionally recognized reason, so he urges impersonal reasons only for his conclusions. Yet his temperament really gives him a stronger bias than any of his more strictly objective premises. It loads the evidence for him one way or the other, making for a more sentimental or a more hard-hearted view of the universe, just as this fact or that principle would. He trusts his temperament. Wanting a universe that suits it, he believes in any representation of the universe that does suit it. He feels men of opposite temper to be out of key with the world's character, and in his heart considers them incompetent and `not in it,' in the philosophic business, even tho they may far excel him in dialectical ability.

\wjparagraph
Yet in the forum he can make no claim, on the bare ground of his temperament, to superior discernment or authority. There arises thus a certain insincerity in our philosophic discussions: the potentest of all our premises is never mentioned. I am sure it would contribute to clearness if in these lectures we should break this rule and mention it, and I accordingly feel free to do so.

\wjparagraph
Of course I am talking here of very positively marked men, men of radical idiosyncracy, who have set their stamp and likeness on philosophy and figure in its history. Plato, Locke, Hegel, Spencer, are such temperamental thinkers. Most of us have, of course, no very definite intellectual temperament, we are a mixture of opposite ingredients, each one present very moderately. We hardly know our own preferences in abstract matters; some of us are easily talked out of them, and end by following the fashion or taking up with the beliefs of the most impressive philosopher in our neighborhood, whoever he may be. But the one thing that has \emph{counted} so far in philosophy is that a man should see things, see them straight in his own peculiar way, and be dissatisfied with any opposite way of seeing them. There is no reason to suppose that this strong temperamental vision is from now onward to count no longer in the history of man's beliefs.

\wjparagraph
Now the particular difference of temperament that I have in mind in making these remarks is one that has counted in literature, art, government and manners as well as in philosophy. In manners we find formalists and free-and-easy persons. In government, authoritarians and anarchists. In literature, purists or academicals, and realists. In art, classics and romantics. You recognize these contrasts as familiar; well, in philosophy we have a very similar contrast expressed in the pair of terms `rationalist' and `empiricist,' `empiricist' meaning your lover of facts in all their crude variety, `rationalist' meaning your devotee to abstract and eternal principles. No one can live an hour without both facts and principles, so it is a difference rather of emphasis; yet it breeds antipathies of the most pungent character between those who lay the emphasis differently; and we shall find it extraordinarily convenient to express a certain contrast in men's ways of taking their universe, by talking of the `empiricist' and of the `rationalist' temper. These terms make the contrast simple and massive.

\wjparagraph
More simple and massive than are usually the men of whom the terms are predicated. For every sort of permutation and combination is possible in human nature; and if I now proceed to define more fully what I have in mind when I speak of rationalists and empiricists, by adding to each of those titles some secondary qualifying characteristics, I beg you to regard my conduct as to a certain extent arbitrary. I select types of combination that nature offers very frequently, but by no means uniformly, and I select them solely for their convenience in helping me to my ulterior purpose of characterizing pragmatism. Historically we find the terms `intellectualism' and `sensationalism' used as synonyms of `rationalism' and `empiricism.' Well, nature seems to combine most frequently with intellectualism an idealistic and optimistic tendency. Empiricists on the other hand are not uncommonly materialistic, and their optimism is apt to be decidedly conditional and tremulous. Rationalism is always monistic. It starts from wholes and universals, and makes much of the unity of things. Empiricism starts from the parts, and makes of the whole a collection-is not averse therefore to calling itself pluralistic. Rationalism usually considers itself more religious than empiricism, but there is much to say about this claim, so I merely mention it. It is a true claim when the individual rationalist is what is called a man of feeling, and when the individual empiricist prides himself on being hard-headed. In that case the rationalist will usually also be in favor of what is called free-will, and the empiricist will be a fatalist--- I use the terms most popularly current. The rationalist finally will be of dogmatic temper in his affirmations, while the empiricist may be more sceptical and open to discussion.

\wjparagraph
I will write these traits down in two columns. I think you will practically recognize the two types of mental make-up that I mean if I head the columns by the titles `tender-minded' and `tough-minded' respectively.

\bigskip

\noindent \framebox{\begin{minipage}{.45\linewidth} \textsc{The Tender-Minded}\par Rationalistic (going by `principles')\\ Intellectualistic\\ Idealistic\\ Optimistic\\ Religious\\ Free-willist\\ Monistic\\ Dogmatical \end{minipage}}\hfill \framebox{\begin{minipage}{.45\linewidth} \textsc{The Tough-Minded}\par Empiricist (going by `facts')\\ Sensationalistic\\ Materialistic\\ Pessimistic\\ Irreligious\\ Fatalistic\\ Pluralistic\\ Sceptical \end{minipage}}

\bigskip

\wjparagraph
Pray postpone for a moment the question whether the two contrasted mixtures which I have written down are each inwardly coherent and self-consistent or not--- I shall very soon have a good deal to say on that point. It suffices for our immediate purpose that tender-minded and tough-minded people, characterized as I have written them down, do both exist. Each of you probably knows some well-marked example of each type, and you know what each example thinks of the example on the other side of the line. They have a low opinion of each other. Their antagonism, whenever as individuals their temperaments have been intense, has formed in all ages a part of the philosophic atmosphere of the time. It forms a part of the philosophic atmosphere to-day. The tough think of the tender as sentimentalists and soft-heads. The tender feel the tough to be unrefined, callous, or brutal. Their mutual reaction is very much like that that takes place when Bostonian tourists mingle with a population like that of Cripple Creek. Each type believes the other to be inferior to itself; but disdain in the one case is mingled with amusement, in the other it has a dash of fear.

\wjparagraph
Now, as I have already insisted, few of us are tender-foot Bostonians pure and simple, and few are typical Rocky Mountain toughs, in philosophy. Most of us have a hankering for the good things on both sides of the line. Facts are good, of course--- give us lots of facts. Principles are good--- give us plenty of principles. The world is indubitably one if you look at it in one way, but as indubitably is it many, if you look at it in another. It is both one and many--- let us adopt a sort of pluralistic monism. Everything of course is necessarily determined, and yet of course our wills are free: a sort of free-will determinism is the true philosophy. The evil of the parts is undeniable; but the whole can't be evil: so practical pessimism may be combined with metaphysical optimism. And so forth--- your ordinary philosophic layman never being a radical, never straightening out his system, but living vaguely in one plausible compartment of it or another to suit the temptations of successive hours.

\wjparagraph
But some of us are more than mere laymen in philosophy. We are worthy of the name of amateur athletes, and are vexed by too much inconsistency and vacillation in our creed. We cannot preserve a good intellectual conscience so long as we keep mixing incompatibles from opposite sides of the line.

\wjparagraph
And now I come to the first positively important point which I wish to make. Never were as many men of a decidedly empiricist proclivity in existence as there are at the present day. Our children, one may say, are almost born scientific. But our esteem for facts has not neutralized in us all religiousness. It is itself almost religious. Our scientific temper is devout. Now take a man of this type, and let him be also a philosophic amateur, unwilling to mix a hodge-podge system after the fashion of a common layman, and what does he find his situation to be, in this blessed year of our Lord 1906? He wants facts; he wants science; but he also wants a religion. And being an amateur and not an independent originator in philosophy he naturally looks for guidance to the experts and professionals whom he finds already in the field. A very large number of you here present, possibly a majority of you, are amateurs of just this sort.

\wjparagraph
Now what kinds of philosophy do you find actually offered to meet your need? You find an empirical philosophy that is not religious enough, and a religious philosophy that is not empirical enough for your purpose. If you look to the quarter where facts are most considered you find the whole tough-minded program in operation, and the `conflict between science and religion' in full blast. Either it is that Rocky Mountain tough of a Haeckel with his materialistic monism, his ether-god and his jest at your God as a `gaseous vertebrate'; or it is Spencer treating the world's history as a redistribution of matter and motion solely, and bowing religion politely out at the front door:--- she may indeed continue to exist, but she must never show her face inside the temple. For a hundred and fifty years past the progress of science has seemed to mean the enlargement of the material universe and the diminution of man's importance. The result is what one may call the growth of naturalistic or positivistic feeling. Man is no law-giver to nature, he is an absorber. She it is who stands firm; he it is who must accommodate himself. Let him record truth, inhuman tho it be, and submit to it! The romantic spontaneity and courage are gone, the vision is materialistic and depressing. Ideals appear as inert by-products of physiology; what is higher is explained by what is lower and treated forever as a case of `nothing but'--- nothing but something else of a quite inferior sort. You get, in short, a materialistic universe, in which only the tough-minded find themselves congenially at home.

\wjparagraph
If now, on the other hand, you turn to the religious quarter for consolation, and take counsel of the tender-minded philosophies, what do you find?

\wjparagraph
Religious philosophy in our day and generation is, among us English-reading people, of two main types. One of these is more radical and aggressive, the other has more the air of fighting a slow retreat. By the more radical wing of religious philosophy I mean the so-called transcendental idealism of the Anglo-Hegelian school, the philosophy of such men as Green, the Cairds, Bosanquet, and Royce. This philosophy has greatly influenced the more studious members of our protestant ministry. It is pantheistic, and undoubtedly it has already blunted the edge of the traditional theism in protestantism at large.

\wjparagraph
That theism remains, however. It is the lineal descendant, through one stage of concession after another, of the dogmatic scholastic theism still taught rigorously in the seminaries of the catholic church. For a long time it used to be called among us the philosophy of the Scottish school. It is what I meant by the philosophy that has the air of fighting a slow retreat. Between the encroachments of the hegelians and other philosophers of the `Absolute,' on the one hand, and those of the scientific evolutionists and agnostics, on the other, the men that give us this kind of a philosophy, James Martineau, Professor Bowne, Professor Ladd and others, must feel themselves rather tightly squeezed. Fair-minded and candid as you like, this philosophy is not radical in temper. It is eclectic, a thing of compromises, that seeks a \emph{modus vivendi} above all things. It accepts the facts of Darwinism, the facts of cerebral physiology, but it does nothing active or enthusiastic with them. It lacks the victorious and aggressive note. It lacks \emph{prestige} in consequence; whereas absolutism has a certain \emph{prestige} due to the more radical style of it.

\wjparagraph
These two systems are what you have to choose between if you turn to the tender-minded school. And if you are the lovers of facts I have supposed you to be, you find the trail of the serpent of rationalism, of intellectualism, over everything that lies on that side of the line. You escape indeed the materialism that goes with the reigning empiricism; but you pay for your escape by losing contact with the concrete parts of life. The more absolutistic philosophers dwell on so high a level of abstraction that they never even try to come down. The absolute mind which they offer us, the mind that makes our universe by thinking it, might, for aught they show us to the contrary, have made any one of a million other universes just as well as this. You can deduce no single actual particular from the notion of it. It is compatible with any state of things whatever being true here below. And the theistic God is almost as sterile a principle. You have to go to the world which he has created to get any inkling of his actual character: he is the kind of god that has once for all made that kind of a world. The God of the theistic writers lives on as purely abstract heights as does the Absolute. Absolutism has a certain sweep and dash about it, while the usual theism is more insipid, but both are equally remote and vacuous. What you want is a philosophy that will not only exercise your powers of intellectual abstraction, but that will make some positive connexion with this actual world of finite human lives.

\wjparagraph
You want a system that will combine both things, the scientific loyalty to facts and willingness to take account of them, the spirit of adaptation and accommodation, in short, but also the old confidence in human values and the resultant spontaneity, whether of the religious or of the romantic type. And this is then your dilemma: you find the two parts of your quaesitum hopelessly separated. You find empiricism with inhumanism and irreligion; or else you find a rationalistic philosophy that indeed may call itself religious, but that keeps out of all definite touch with concrete facts and joys and sorrows.

\wjparagraph
I am not sure how many of you live close enough to philosophy to realize fully what I mean by this last reproach, so I will dwell a little longer on that unreality in all rationalistic systems by which your serious believer in facts is so apt to feel repelled.

\wjparagraph
I wish that I had saved the first couple of pages of a thesis which a student handed me a year or two ago. They illustrated my point so clearly that I am sorry I cannot read them to you now. This young man, who was a graduate of some Western college, began by saying that he had always taken for granted that when you entered a philosophic class-room you had to open relations with a universe entirely distinct from the one you left behind you in the street. The two were supposed, he said, to have so little to do with each other, that you could not possibly occupy your mind with them at the same time. The world of concrete personal experiences to which the street belongs is multitudinous beyond imagination, tangled, muddy, painful and perplexed. The world to which your philosophy-professor introduces you is simple, clean and noble. The contradictions of real life are absent from it. Its architecture is classic. Principles of reason trace its outlines, logical necessities cement its parts. Purity and dignity are what it most expresses. It is a kind of marble temple shining on a hill.

\wjparagraph
In point of fact it is far less an account of this actual world than a clear addition built upon it, a classic sanctuary in which the rationalist fancy may take refuge from the intolerably confused and gothic character which mere facts present. It is no \emph{explanation} of our concrete universe, it is another thing altogether, a substitute for it, a remedy, a way of escape.

\wjparagraph
Its temperament, if I may use the word temperament here, is utterly alien to the temperament of existence in the concrete. \emph{Refinement} is what characterizes our intellectualist philosophies. They exquisitely satisfy that craving for a refined object of contemplation which is so powerful an appetite of the mind. But I ask you in all seriousness to look abroad on this colossal universe of concrete facts, on their awful bewilderments, their surprises and cruelties, on the wildness which they show, and then to tell me whether `refined' is the one inevitable descriptive adjective that springs to your lips.

\wjparagraph
Refinement has its place in things, true enough. But a philosophy that breathes out nothing but refinement will never satisfy the empiricist temper of mind. It will seem rather a monument of artificiality. So we find men of science preferring to turn their backs on metaphysics as on something altogether cloistered and spectral, and practical men shaking philosophy's dust off their feet and following the call of the wild.

\wjparagraph
Truly there is something a little ghastly in the satisfaction with which a pure but unreal system will fill a rationalist mind. Leibnitz was a rationalist mind, with infinitely more interest in facts than most rationalist minds can show. Yet if you wish for superficiality incarnate, you have only to read that charmingly written `Theodicee' of his, in which he sought to justify the ways of God to man, and to prove that the world we live in is the best of possible worlds. Let me quote a specimen of what I mean.

\wjparagraph
Among other obstacles to his optimistic philosophy, it falls to Leibnitz to consider the number of the eternally damned. That it is infinitely greater, in our human case, than that of those saved he assumes as a premise from the theologians, and then proceeds to argue in this way. Even then, he says:

\wjparagraph
``The evil will appear as almost nothing in comparison with the good, if we once consider the real magnitude of the City of God. Coelius Secundus Curio has written a little book, \emph{De Amplitudine Regni Coelestis}, which was reprinted not long ago. But he failed to compass the extent of the kingdom of the heavens. The ancients had small ideas of the works of God. ... It seemed to them that only our earth had inhabitants, and even the notion of our antipodes gave them pause. The rest of the world for them consisted of some shining globes and a few crystalline spheres. But to-day, whatever be the limits that we may grant or refuse to the Universe we must recognize in it a countless number of globes, as big as ours or bigger, which have just as much right as it has to support rational inhabitants, tho it does not follow that these need all be men. Our earth is only one among the six principal satellites of our sun. As all the fixed stars are suns, one sees how small a place among visible things our earth takes up, since it is only a satellite of one among them. Now all these suns \emph{may} be inhabited by none but happy creatures; and nothing obliges us to believe that the number of damned persons is very great; for a \emph{very few instances and samples suffice for the utility which good draws from evil}. Moreover, since there is no reason to suppose that there are stars everywhere, may there not be a great space beyond the region of the stars? And this immense space, surrounding all this region, ... may be replete with happiness and glory. ... What now becomes of the consideration of our Earth and of its denizens? Does it not dwindle to something incomparably less than a physical point, since our Earth is but a point compared with the distance of the fixed stars. Thus the part of the Universe which we know, being almost lost in nothingness compared with that which is unknown to us, but which we are yet obliged to admit; and all the evils that we know lying in this almost-nothing; it follows that the evils may be almost-nothing in comparison with the goods that the Universe contains.''

\wjparagraph
Leibnitz continues elsewhere: ``There is a kind of justice which aims neither at the amendment of the criminal, nor at furnishing an example to others, nor at the reparation of the injury. This justice is founded in pure fitness, which finds a certain satisfaction in the expiation of a wicked deed. The Socinians and Hobbes objected to this punitive justice, which is properly vindictive justice and which God has reserved for himself at many junctures. ... It is always founded in the fitness of things, and satisfies not only the offended party, but all wise lookers-on, even as beautiful music or a fine piece of architecture satisfies a well-constituted mind. It is thus that the torments of the damned continue, even tho they serve no longer to turn anyone away from sin, and that the rewards of the blest continue, even tho they confirm no one in good ways. The damned draw to themselves ever new penalties by their continuing sins, and the blest attract ever fresh joys by their unceasing progress in good. Both facts are founded on the principle of fitness, ... for God has made all things harmonious in perfection as I have already said.''

\wjparagraph
Leibnitz's feeble grasp of reality is too obvious to need comment from me. It is evident that no realistic image of the experience of a damned soul had ever approached the portals of his mind. Nor had it occurred to him that the smaller is the number of `samples' of the genus `lost-soul' whom God throws as a sop to the eternal fitness, the more unequitably grounded is the glory of the blest. What he gives us is a cold literary exercise, whose cheerful substance even hell-fire does not warm.

\wjparagraph
And do not tell me that to show the shallowness of rationalist philosophizing I have had to go back to a shallow wigpated age. The optimism of present-day rationalism sounds just as shallow to the fact-loving mind. The actual universe is a thing wide open, but rationalism makes systems, and systems must be closed. For men in practical life perfection is something far off and still in process of achievement. This for rationalism is but the illusion of the finite and relative: the absolute ground of things is a perfection eternally complete.

\wjparagraph
I find a fine example of revolt against the airy and shallow optimism of current religious philosophy in a publication of that valiant anarchistic writer Morrison I. Swift. Mr. Swift's anarchism goes a little farther than mine does, but I confess that I sympathize a good deal, and some of you, I know, will sympathize heartily with his dissatisfaction with the idealistic optimisms now in vogue. He begins his pamphlet on \emph{Human Submission} with a series of city reporter's items from newspapers (suicides, deaths from starvation, and the like) as specimens of our civilized regime. For instance:

\wjparagraph
`` `After trudging through the snow from one end of the city to the other in the vain hope of securing employment, and with his wife and six children without food and ordered to leave their home in an upper east side tenement house because of non-payment of rent, John Corcoran, a clerk, to-day ended his life by drinking carbolic acid. Corcoran lost his position three weeks ago through illness, and during the period of idleness his scanty savings disappeared. Yesterday he obtained work with a gang of city snow shovelers, but he was too weak from illness and was forced to quit after an hour's trial with the shovel. Then the weary task of looking for employment was again resumed. Thoroughly discouraged, Corcoran returned to his home late last night to find his wife and children without food and the notice of dispossession on the door.' On the following morning he drank the poison.

\wjparagraph
``The records of many more such cases lie before me [Mr. Swift goes on]; an encyclopedia might easily be filled with their kind. These few I cite as an interpretation of the universe. `We are aware of the presence of God in His world,' says a writer in a recent English Review. [The very presence of ill in the temporal order is the condition of the perfection of the eternal order, writes Professor Royce (\emph{The World and the Individual}, II, 385).] `The Absolute is the richer for every discord, and for all diversity which it embraces,' says F. H. Bradley (Appearance and Reality, 204). He means that these slain men make the universe richer, and that is Philosophy. But while Professors Royce and Bradley and a whole host of guileless thoroughfed thinkers are unveiling Reality and the Absolute and explaining away evil and pain, this is the condition of the only beings known to us anywhere in the universe with a developed consciousness of what the universe is. What these people experience \emph{is} Reality. It gives us an absolute phase of the universe. It is the personal experience of those most qualified in all our circle of knowledge to \emph{have} experience, to tell us \emph{what} is. Now, what does \emph{thinking about} the experience of these persons come to compared with directly, personally feeling it, as they feel it? The philosophers are dealing in shades, while those who live and feel know truth. And the mind of mankind--- not yet the mind of philosophers and of the proprietary class--- but of the great mass of the silently thinking and feeling men, is coming to this view. They are judging the universe as they have heretofore permitted the hierophants of religion and learning to judge \emph{them}. ...

\wjparagraph
``This Cleveland workingman, killing his children and himself [another of the cited cases], is one of the elemental, stupendous facts of this modern world and of this universe. It cannot be glozed over or minimized away by all the treatises on God, and Love, and Being, helplessly existing in their haughty monumental vacuity. This is one of the simple irreducible elements of this world's life after millions of years of divine opportunity and twenty centuries of Christ. It is in the moral world like atoms or sub-atoms in the physical, primary, indestructible. And what it blazons to man is the ... imposture of all philosophy which does not see in such events the consummate factor of conscious experience. These facts invincibly prove religion a nullity. Man will not give religion two thousand centuries or twenty centuries more to try itself and waste human time; its time is up, its probation is ended. Its own record ends it. Mankind has not sons and eternities to spare for trying out discredited systems....''\footnote{Morrison I. Swift, \emph{Human Submission}, Part Second, Philadelphia, Liberty Press, 1905, pp. 4-10.}

\wjparagraph
Such is the reaction of an empiricist mind upon the rationalist bill of fare. It is an absolute `No, I thank you.' ``Religion,'' says Mr. Swift, ``is like a sleep-walker to whom actual things are blank.'' And such, tho possibly less tensely charged with feeling, is the verdict of every seriously inquiring amateur in philosophy to-day who turns to the philosophy-professors for the wherewithal to satisfy the fulness of his nature's needs. Empiricist writers give him a materialism, rationalists give him something religious, but to that religion ``actual things are blank.'' He becomes thus the judge of us philosophers. Tender or tough, he finds us wanting. None of us may treat his verdicts disdainfully, for after all, his is the typically perfect mind, the mind the sum of whose demands is greatest, the mind whose criticisms and dissatisfactions are fatal in the long run.

\wjparagraph
It is at this point that my own solution begins to appear. I offer the oddly-named thing pragmatism as a philosophy that can satisfy both kinds of demand. It can remain religious like the rationalisms, but at the same time, like the empiricisms, it can preserve the richest intimacy with facts. I hope I may be able to leave many of you with as favorable an opinion of it as I preserve myself. Yet, as I am near the end of my hour, I will not introduce pragmatism bodily now. I will begin with it on the stroke of the clock next time. I prefer at the present moment to return a little on what I have said.

\wjparagraph
If any of you here are professional philosophers, and some of you I know to be such, you will doubtless have felt my discourse so far to have been crude in an unpardonable, nay, in an almost incredible degree. Tender-minded and tough-minded, what a barbaric disjunction! And, in general, when philosophy is all compacted of delicate intellectualities and subtleties and scrupulosities, and when every possible sort of combination and transition obtains within its bounds, what a brutal caricature and reduction of highest things to the lowest possible expression is it to represent its field of conflict as a sort of rough-and-tumble fight between two hostile temperaments! What a childishly external view! And again, how stupid it is to treat the abstractness of rationalist systems as a crime, and to damn them because they offer themselves as sanctuaries and places of escape, rather than as prolongations of the world of facts. Are not all our theories just remedies and places of escape? And, if philosophy is to be religious, how can she be anything else than a place of escape from the crassness of reality's surface? What better thing can she do than raise us out of our animal senses and show us another and a nobler home for our minds in that great framework of ideal principles subtending all reality, which the intellect divines? How can principles and general views ever be anything but abstract outlines? Was Cologne cathedral built without an architect's plan on paper? Is refinement in itself an abomination? Is concrete rudeness the only thing that's true?

\wjparagraph
Believe me, I feel the full force of the indictment. The picture I have given is indeed monstrously over-simplified and rude. But like all abstractions, it will prove to have its use. If philosophers can treat the life of the universe abstractly, they must not complain of an abstract treatment of the life of philosophy itself. In point of fact the picture I have given is, however coarse and sketchy, literally true. Temperaments with their cravings and refusals do determine men in their philosophies, and always will. The details of systems may be reasoned out piecemeal, and when the student is working at a system, he may often forget the forest for the single tree. But when the labor is accomplished, the mind always performs its big summarizing act, and the system forthwith stands over against one like a living thing, with that strange simple note of individuality which haunts our memory, like the wraith of the man, when a friend or enemy of ours is dead.

\wjparagraph
Not only Walt Whitman could write ``who touches this book touches a man.'' The books of all the great philosophers are like so many men. Our sense of an essential personal flavor in each one of them, typical but indescribable, is the finest fruit of our own accomplished philosophic education. What the system pretends to be is a picture of the great universe of God. What it is--- and oh so flagrantly!--- is the revelation of how intensely odd the personal flavor of some fellow creature is. Once reduced to these terms (and all our philosophies get reduced to them in minds made critical by learning) our commerce with the systems reverts to the informal, to the instinctive human reaction of satisfaction or dislike. We grow as peremptory in our rejection or admission, as when a person presents himself as a candidate for our favor; our verdicts are couched in as simple adjectives of praise or dispraise. We measure the total character of the universe as we feel it, against the flavor of the philosophy proffered us, and one word is enough.

\wjparagraph
``Statt der lebendigen Natur,'' we say, ``da Gott die Menschen schuf hinein,''\footnote{[James is quoting Goethe's \emph{Faust}. In English, the line is ``Instead of living nature in which God created man''--- beyond the lines quoted here, it continues--- ``you're surrounded by smoke and rot, animals' skeletons and dead men's bones.'']}--- that nebulous concoction, that wooden, that straight-laced thing, that crabbed artificiality, that musty schoolroom product, that sick man's dream! Away with it. Away with all of them! Impossible! Impossible!

\wjparagraph
Our work over the details of his system is indeed what gives us our resultant impression of the philosopher, but it is on the resultant impression itself that we react. Expertness in philosophy is measured by the definiteness of our summarizing reactions, by the immediate perceptive epithet with which the expert hits such complex objects off. But great expertness is not necessary for the epithet to come. Few people have definitely articulated philosophies of their own. But almost everyone has his own peculiar sense of a certain total character in the universe, and of the inadequacy fully to match it of the peculiar systems that he knows. They don't just cover \emph{his} world. One will be too dapper, another too pedantic, a third too much of a job-lot of opinions, a fourth too morbid, and a fifth too artificial, or what not. At any rate he and we know offhand that such philosophies are out of plumb and out of key and out of `whack,' and have no business to speak up in the universe's name. Plato, Locke, Spinoza, Mill, Caird, Hegel--- I prudently avoid names nearer home!--- I am sure that to many of you, my hearers, these names are little more than reminders of as many curious personal ways of falling short. It would be an obvious absurdity if such ways of taking the universe were actually true. We philosophers have to reckon with such feelings on your part. In the last resort, I repeat, it will be by them that all our philosophies shall ultimately be judged. The finally victorious way of looking at things will be the most completely \emph{impressive} way to the normal run of minds.

\wjparagraph
One word more--- namely about philosophies necessarily being abstract outlines. There are outlines and outlines, outlines of buildings that are \emph{fat}, conceived in the cube by their planner, and outlines of buildings invented flat on paper, with the aid of ruler and compass. These remain skinny and emaciated even when set up in stone and mortar, and the outline already suggests that result. An outline in itself is meagre, truly, but it does not necessarily suggest a meagre thing. It is the essential meagreness of \emph{what is suggested} by the usual rationalistic philosophies that moves empiricists to their gesture of rejection. The case of Herbert Spencer's system is much to the point here. Rationalists feel his fearful array of insufficiencies. His dry schoolmaster temperament, the hurdy-gurdy monotony of him, his preference for cheap makeshifts in argument, his lack of education even in mechanical principles, and in general the vagueness of all his fundamental ideas, his whole system wooden, as if knocked together out of cracked hemlock boards--- and yet the half of England wants to bury him in Westminster Abbey.

\wjparagraph
Why? Why does Spencer call out so much reverence in spite of his weakness in rationalistic eyes? Why should so many educated men who feel that weakness, you and I perhaps, wish to see him in the Abbey notwithstanding?

\wjparagraph
Simply because we feel his heart to be \emph{in the right place} philosophically. His principles may be all skin and bone, but at any rate his books try to mould themselves upon the particular shape of this, particular world's carcase. The noise of facts resounds through all his chapters, the citations of fact never cease, he emphasizes facts, turns his face towards their quarter; and that is enough. It means the right \emph{kind} of thing for the empiricist mind.

\wjparagraph
The pragmatistic philosophy of which I hope to begin talking in my next lecture preserves as cordial a relation with facts, and, unlike Spencer's philosophy, it neither begins nor ends by turning positive religious constructions out of doors--- it treats them cordially as well.

\wjparagraph
I hope I may lead you to find it just the mediating way of thinking that you require. 


\section*{Lecture II: What Pragmatism Means}
\setcounter{wjparagraph}{0}

\wjparagraph
Some years ago, being with a camping party in the mountains, I returned from a solitary ramble to find everyone engaged in a ferocious metaphysical dispute. The corpus of the dispute was a squirrel--- a live squirrel supposed to be clinging to one side of a tree-trunk; while over against the tree's opposite side a human being was imagined to stand. This human witness tries to get sight of the squirrel by moving rapidly round the tree, but no matter how fast he goes, the squirrel moves as fast in the opposite direction, and always keeps the tree between himself and the man, so that never a glimpse of him is caught. The resultant metaphysical problem now is this: \emph{Does the man go round the squirrel or not?} He goes round the tree, sure enough, and the squirrel is on the tree; but does he go round the squirrel? In the unlimited leisure of the wilderness, discussion had been worn threadbare. Everyone had taken sides, and was obstinate; and the numbers on both sides were even. Each side, when I appeared, therefore appealed to me to make it a majority. Mindful of the scholastic adage that whenever you meet a contradiction you must make a distinction, I immediately sought and found one, as follows: ``Which party is right,'' I said, ``depends on what you \emph{practically mean} by `going round' the squirrel. If you mean passing from the north of him to the east, then to the south, then to the west, and then to the north of him again, obviously the man does go round him, for he occupies these successive positions. But if on the contrary you mean being first in front of him, then on the right of him, then behind him, then on his left, and finally in front again, it is quite as obvious that the man fails to go round him, for by the compensating movements the squirrel makes, he keeps his belly turned towards the man all the time, and his back turned away. Make the distinction, and there is no occasion for any farther dispute. You are both right and both wrong according as you conceive the verb `to go round' in one practical fashion or the other.''

\wjparagraph
Altho one or two of the hotter disputants called my speech a shuffling evasion, saying they wanted no quibbling or scholastic hair-splitting, but meant just plain honest English `round,' the majority seemed to think that the distinction had assuaged the dispute.

\wjparagraph
I tell this trivial anecdote because it is a peculiarly simple example of what I wish now to speak of as \emph{the pragmatic method}. The pragmatic method is primarily a method of settling metaphysical disputes that otherwise might be interminable. Is the world one or many?--- fated or free?--- material or spiritual?--- here are notions either of which may or may not hold good of the world; and disputes over such notions are unending. The pragmatic method in such cases is to try to interpret each notion by tracing its respective practical consequences. What difference would it practically make to anyone if this notion rather than that notion were true? If no practical difference whatever can be traced, then the alternatives mean practically the same thing, and all dispute is idle. Whenever a dispute is serious, we ought to be able to show some practical difference that must follow from one side or the other's being right.

\wjparagraph
A glance at the history of the idea will show you still better what pragmatism means. The term is derived from the same Greek word $\pi\rho\alpha\gamma\mu\alpha$, meaning action, from which our words `practice' and `practical' come. It was first introduced into philosophy by Mr. Charles Peirce in 1878. In an article entitled `How to Make Our Ideas Clear,' in the `Popular Science Monthly' for January of that year\footnote{Translated in the \emph{Revue Philosophique} for January, 1879 (vol. vii).} Mr. Peirce, after pointing out that our beliefs are really rules for action, said that to develop a thought's meaning, we need only determine what conduct it is fitted to produce: that conduct is for us its sole significance. And the tangible fact at the root of all our thought-distinctions, however subtle, is that there is no one of them so fine as to consist in anything but a possible difference of practice. To attain perfect clearness in our thoughts of an object, then, we need only consider what conceivable effects of a practical kind the object may involve--- what sensations we are to expect from it, and what reactions we must prepare. Our conception of these effects, whether immediate or remote, is then for us the whole of our conception of the object, so far as that conception has positive significance at all.

\wjparagraph
This is the principle of Peirce, the principle of pragmatism. It lay entirely unnoticed by anyone for twenty years, until I, in an address before Professor Howison's philosophical union at the university of California, brought it forward again and made a special application of it to religion. By that date (1898) the times seemed ripe for its reception. The word `pragmatism' spread, and at present it fairly spots the pages of the philosophic journals. On all hands we find the `pragmatic movement' spoken of, sometimes with respect, sometimes with contumely, seldom with clear understanding. It is evident that the term applies itself conveniently to a number of tendencies that hitherto have lacked a collective name, and that it has `come to stay.'

\wjparagraph
To take in the importance of Peirce's principle, one must get accustomed to applying it to concrete cases. I found a few years ago that Ostwald, the illustrious Leipzig chemist, had been making perfectly distinct use of the principle of pragmatism in his lectures on the philosophy of science, tho he had not called it by that name.

\wjparagraph
``All realities influence our practice,'' he wrote me, ``and that influence is their meaning for us. I am accustomed to put questions to my classes in this way: In what respects would the world be different if this alternative or that were true? If I can find nothing that would become different, then the alternative has no sense.''

\wjparagraph
That is, the rival views mean practically the same thing, and meaning, other than practical, there is for us none. Ostwald in a published lecture gives this example of what he means. Chemists have long wrangled over the inner constitution of certain bodies called `tautomerous.' Their properties seemed equally consistent with the notion that an instable hydrogen atom oscillates inside of them, or that they are instable mixtures of two bodies. Controversy raged; but never was decided. ``It would never have begun,'' says Ostwald, ``if the combatants had asked themselves what particular experimental fact could have been made different by one or the other view being correct. For it would then have appeared that no difference of fact could possibly ensue; and the quarrel was as unreal as if, theorizing in primitive times about the raising of dough by yeast, one party should have invoked a `brownie,' while another insisted on an `elf' as the true cause of the phenomenon.''\footnote{`Theorie und Praxis,' Zeitsch. des Oesterreichischen Ingenieur u. Architecten-Vereines, 1905, Nr. 4 u. 6. I find a still more radical pragmatism than Ostwald's in an address by Professor W. S. Franklin: ``I think that the sickliest notion of physics, even if a student gets it, is that it is `the science of masses, molecules and the ether.' And I think that the healthiest notion, even if a student does not wholly get it, is that physics is the science of the ways of taking hold of bodies and pushing them!'' (\emph{Science}, January 2, 1903.)}

\wjparagraph
It is astonishing to see how many philosophical disputes collapse into insignificance the moment you subject them to this simple test of tracing a concrete consequence. There can \emph{be} no difference anywhere that doesn't \emph{make} a difference elsewhere--- no difference in abstract truth that doesn't express itself in a difference in concrete fact and in conduct consequent upon that fact, imposed on somebody, somehow, somewhere and somewhen. The whole function of philosophy ought to be to find out what definite difference it will make to you and me, at definite instants of our life, if this world-formula or that world-formula be the true one.

\wjparagraph
There is absolutely nothing new in the pragmatic method. Socrates was an adept at it. Aristotle used it methodically. Locke, Berkeley and Hume made momentous contributions to truth by its means. Shadworth Hodgson keeps insisting that realities are only what they are `known-as.' But these forerunners of pragmatism used it in fragments: they were preluders only. Not until in our time has it generalized itself, become conscious of a universal mission, pretended to a conquering destiny. I believe in that destiny, and I hope I may end by inspiring you with my belief.

\wjparagraph
Pragmatism represents a perfectly familiar attitude in philosophy, the empiricist attitude, but it represents it, as it seems to me, both in a more radical and in a less objectionable form than it has ever yet assumed. A pragmatist turns his back resolutely and once for all upon a lot of inveterate habits dear to professional philosophers. He turns away from abstraction and insufficiency, from verbal solutions, from bad a priori reasons, from fixed principles, closed systems, and pretended absolutes and origins. He turns towards concreteness and adequacy, towards facts, towards action, and towards power. That means the empiricist temper regnant, and the rationalist temper sincerely given up. It means the open air and possibilities of nature, as against dogma, artificiality and the pretence of finality in truth.

\wjparagraph
At the same time it does not stand for any special results. It is a method only. But the general triumph of that method would mean an enormous change in what I called in my last lecture the `temperament' of philosophy. Teachers of the ultra-rationalistic type would be frozen out, much as the courtier type is frozen out in republics, as the ultramontane type of priest is frozen out in protestant lands. Science and metaphysics would come much nearer together, would in fact work absolutely hand in hand.

\wjparagraph
Metaphysics has usually followed a very primitive kind of quest. You know how men have always hankered after unlawful magic, and you know what a great part, in magic, \emph{words} have always played. If you have his name, or the formula of incantation that binds him, you can control the spirit, genie, afrite, or whatever the power may be. Solomon knew the names of all the spirits, and having their names, he held them subject to his will. So the universe has always appeared to the natural mind as a kind of enigma, of which the key must be sought in the shape of some illuminating or power-bringing word or name. That word names the universe's \emph{principle}, and to possess it is, after a fashion, to possess the universe itself. `God,' `Matter,' `Reason,' `the Absolute,' `Energy,' are so many solving names. You can rest when you have them. You are at the end of your metaphysical quest.

\wjparagraph
But if you follow the pragmatic method, you cannot look on any such word as closing your quest. You must bring out of each word its practical cash-value, set it at work within the stream of your experience. It appears less as a solution, then, than as a program for more work, and more particularly as an indication of the ways in which existing realities may be \emph{changed}.

\wjparagraph
\emph{Theories thus become instruments, not answers to enigmas, in which we can rest.} We don't lie back upon them, we move forward, and, on occasion, make nature over again by their aid. Pragmatism unstiffens all our theories, limbers them up and sets each one at work. Being nothing essentially new, it harmonizes with many ancient philosophic tendencies. It agrees with nominalism for instance, in always appealing to particulars; with utilitarianism in emphasizing practical aspects; with positivism in its disdain for verbal solutions, useless questions, and metaphysical abstractions.

\wjparagraph
All these, you see, are \emph{anti-intellectualist} tendencies. Against rationalism as a pretension and a method, pragmatism is fully armed and militant. But, at the outset, at least, it stands for no particular results. It has no dogmas, and no doctrines save its method. As the young Italian pragmatist Papini has well said, it lies in the midst of our theories, like a corridor in a hotel. Innumerable chambers open out of it. In one you may find a man writing an atheistic volume; in the next someone on his knees praying for faith and strength; in a third a chemist investigating a body's properties. In a fourth a system of idealistic metaphysics is being excogitated; in a fifth the impossibility of metaphysics is being shown. But they all own the corridor, and all must pass through it if they want a practicable way of getting into or out of their respective rooms.

\wjparagraph
No particular results then, so far, but only an attitude of orientation, is what the pragmatic method means. \emph{The attitude of looking away from first things, principles, `categories,' supposed necessities; and of looking towards last things, fruits, consequences, facts.}

\wjparagraph
So much for the pragmatic method! You may say that I have been praising it rather than explaining it to you, but I shall presently explain it abundantly enough by showing how it works on some familiar problems. Meanwhile the word pragmatism has come to be used in a still wider sense, as meaning also a certain theory of \emph{truth}. I mean to give a whole lecture to the statement of that theory, after first paving the way, so I can be very brief now. But brevity is hard to follow, so I ask for your redoubled attention for a quarter of an hour. If much remains obscure, I hope to make it clearer in the later lectures.

\wjparagraph
One of the most successfully cultivated branches of philosophy in our time is what is called inductive logic, the study of the conditions under which our sciences have evolved. Writers on this subject have begun to show a singular unanimity as to what the laws of nature and elements of fact mean, when formulated by mathematicians, physicists and chemists. When the first mathematical, logical and natural uniformities, the first \emph{laws}, were discovered, men were so carried away by the clearness, beauty and simplification that resulted, that they believed themselves to have deciphered authentically the eternal thoughts of the Almighty. His mind also thundered and reverberated in syllogisms. He also thought in conic sections, squares and roots and ratios, and geometrized like Euclid. He made Kepler's laws for the planets to follow; he made velocity increase proportionally to the time in falling bodies; he made the law of the sines for light to obey when refracted; he established the classes, orders, families and genera of plants and animals, and fixed the distances between them. He thought the archetypes of all things, and devised their variations; and when we rediscover any one of these his wondrous institutions, we seize his mind in its very literal intention.

\wjparagraph
But as the sciences have developed farther, the notion has gained ground that most, perhaps all, of our laws are only approximations. The laws themselves, moreover, have grown so numerous that there is no counting them; and so many rival formulations are proposed in all the branches of science that investigators have become accustomed to the notion that no theory is absolutely a transcript of reality, but that any one of them may from some point of view be useful. Their great use is to summarize old facts and to lead to new ones. They are only a man-made language, a conceptual shorthand, as someone calls them, in which we write our reports of nature; and languages, as is well known, tolerate much choice of expression and many dialects.

\wjparagraph
Thus human arbitrariness has driven divine necessity from scientific logic. If I mention the names of Sigwart, Mach, Ostwald, Pearson, Milhaud, Poincare, Duhem, Ruyssen, those of you who are students will easily identify the tendency I speak of, and will think of additional names.

\wjparagraph
Riding now on the front of this wave of scientific logic Messrs. Schiller and Dewey appear with their pragmatistic account of what truth everywhere signifies. Everywhere, these teachers say, `truth' in our ideas and beliefs means the same thing that it means in science. It means, they say, nothing but this, \emph{that ideas (which themselves are but parts of our experience) become true just in so far as they help us to get into satisfactory relation with other parts of our experience}, to summarize them and get about among them by conceptual short-cuts instead of following the interminable succession of particular phenomena. Any idea upon which we can ride, so to speak; any idea that will carry us prosperously from any one part of our experience to any other part, linking things satisfactorily, working securely, simplifying, saving labor; is true for just so much, true in so far forth, true \emph{instrumentally}. This is the `instrumental' view of truth taught so successfully at Chicago, the view that truth in our ideas means their power to `work,' promulgated so brilliantly at Oxford.

\wjparagraph
Messrs. Dewey, Schiller and their allies, in reaching this general conception of all truth, have only followed the example of geologists, biologists and philologists. In the establishment of these other sciences, the successful stroke was always to take some simple process actually observable in operation--- as denudation by weather, say, or variation from parental type, or change of dialect by incorporation of new words and pronunciations--- and then to generalize it, making it apply to all times, and produce great results by summating its effects through the ages.

\wjparagraph
The observable process which Schiller and Dewey particularly singled out for generalization is the familiar one by which any individual settles into \emph{new opinions}. The process here is always the same. The individual has a stock of old opinions already, but he meets a new experience that puts them to a strain. Somebody contradicts them; or in a reflective moment he discovers that they contradict each other; or he hears of facts with which they are incompatible; or desires arise in him which they cease to satisfy. The result is an inward trouble to which his mind till then had been a stranger, and from which he seeks to escape by modifying his previous mass of opinions. He saves as much of it as he can, for in this matter of belief we are all extreme conservatives. So he tries to change first this opinion, and then that (for they resist change very variously), until at last some new idea comes up which he can graft upon the ancient stock with a minimum of disturbance of the latter, some idea that mediates between the stock and the new experience and runs them into one another most felicitously and expediently.

\wjparagraph
This new idea is then adopted as the true one. It preserves the older stock of truths with a minimum of modification, stretching them just enough to make them admit the novelty, but conceiving that in ways as familiar as the case leaves possible. An outr\'{e}e explanation, violating all our preconceptions, would never pass for a true account of a novelty. We should scratch round industriously till we found something less excentric. The most violent revolutions in an individual's beliefs leave most of his old order standing. Time and space, cause and effect, nature and history, and one's own biography remain untouched. New truth is always a go-between, a smoother-over of transitions. It marries old opinion to new fact so as ever to show a minimum of jolt, a maximum of continuity. We hold a theory true just in proportion to its success in solving this `problem of maxima and minima.' But success in solving this problem is eminently a matter of approximation. We say this theory solves it on the whole more satisfactorily than that theory; but that means more satisfactorily to ourselves, and individuals will emphasize their points of satisfaction differently. To a certain degree, therefore, everything here is plastic.

\wjparagraph
The point I now urge you to observe particularly is the part played by the older truths. Failure to take account of it is the source of much of the unjust criticism leveled against pragmatism. Their influence is absolutely controlling. Loyalty to them is the first principle--- in most cases it is the only principle; for by far the most usual way of handling phenomena so novel that they would make for a serious rearrangement of our preconceptions is to ignore them altogether, or to abuse those who bear witness for them.

\wjparagraph
You doubtless wish examples of this process of truth's growth, and the only trouble is their superabundance. The simplest case of new truth is of course the mere numerical addition of new kinds of facts, or of new single facts of old kinds, to our experience--- an addition that involves no alteration in the old beliefs. Day follows day, and its contents are simply added. The new contents themselves are not true, they simply \emph{come} and \emph{are}. Truth is what we say about them, and when we say that they have come, truth is satisfied by the plain additive formula.

\wjparagraph
But often the day's contents oblige a rearrangement. If I should now utter piercing shrieks and act like a maniac on this platform, it would make many of you revise your ideas as to the probable worth of my philosophy. `Radium' came the other day as part of the day's content, and seemed for a moment to contradict our ideas of the whole order of nature, that order having come to be identified with what is called the conservation of energy. The mere sight of radium paying heat away indefinitely out of its own pocket seemed to violate that conservation. What to think? If the radiations from it were nothing but an escape of unsuspected `potential' energy, pre-existent inside of the atoms, the principle of conservation would be saved. The discovery of `helium' as the radiation's outcome, opened a way to this belief. So Ramsay's view is generally held to be true, because, altho it extends our old ideas of energy, it causes a minimum of alteration in their nature.

\wjparagraph
I need not multiply instances. A new opinion counts as `true' just in proportion as it gratifies the individual's desire to assimilate the novel in his experience to his beliefs in stock. It must both lean on old truth and grasp new fact; and its success (as I said a moment ago) in doing this, is a matter for the individual's appreciation. When old truth grows, then, by new truth's addition, it is for subjective reasons. We are in the process and obey the reasons. That new idea is truest which performs most felicitously its function of satisfying our double urgency. It makes itself true, gets itself classed as true, by the way it works; grafting itself then upon the ancient body of truth, which thus grows much as a tree grows by the activity of a new layer of cambium.

\wjparagraph
Now Dewey and Schiller proceed to generalize this observation and to apply it to the most ancient parts of truth. They also once were plastic. They also were called true for human reasons. They also mediated between still earlier truths and what in those days were novel observations. Purely objective truth, truth in whose establishment the function of giving human satisfaction in marrying previous parts of experience with newer parts played no role whatever, is nowhere to be found. The reasons why we call things true is the reason why they \emph{are} true, for `to be true' \emph{means} only to perform this marriage-function.

\wjparagraph
The trail of the human serpent is thus over everything. Truth independent; truth that we \emph{find} merely; truth no longer malleable to human need; truth incorrigible, in a word; such truth exists indeed superabundantly--- or is supposed to exist by rationalistically minded thinkers; but then it means only the dead heart of the living tree, and its being there means only that truth also has its paleontology and its `prescription,' and may grow stiff with years of veteran service and petrified in men's regard by sheer antiquity. But how plastic even the oldest truths nevertheless really are has been vividly shown in our day by the transformation of logical and mathematical ideas, a transformation which seems even to be invading physics. The ancient formulas are reinterpreted as special expressions of much wider principles, principles that our ancestors never got a glimpse of in their present shape and formulation.

\wjparagraph
Mr. Schiller still gives to all this view of truth the name of `Humanism,' but, for this doctrine too, the name of pragmatism seems fairly to be in the ascendant, so I will treat it under the name of pragmatism in these lectures.

\wjparagraph
Such then would be the scope of pragmatism--- first, a method; and second, a genetic theory of what is meant by truth. And these two things must be our future topics.

\wjparagraph
What I have said of the theory of truth will, I am sure, have appeared obscure and unsatisfactory to most of you by reason of us brevity. I shall make amends for that hereafter. In a lecture on `common sense' I shall try to show what I mean by truths grown petrified by antiquity. In another lecture I shall expatiate on the idea that our thoughts become true in proportion as they successfully exert their go-between function. In a third I shall show how hard it is to discriminate subjective from objective factors in Truth's development.\footnote{[James is referring to Lectures V--VII.]} You may not follow me wholly in these lectures; and if you do, you may not wholly agree with me. But you will, I know, regard me at least as serious, and treat my effort with respectful consideration.

\wjparagraph
You will probably be surprised to learn, then, that Messrs. Schiller's and Dewey's theories have suffered a hailstorm of contempt and ridicule. All rationalism has risen against them. In influential quarters Mr. Schiller, in particular, has been treated like an impudent schoolboy who deserves a spanking. I should not mention this, but for the fact that it throws so much sidelight upon that rationalistic temper to which I have opposed the temper of pragmatism. Pragmatism is uncomfortable away from facts. Rationalism is comfortable only in the presence of abstractions. This pragmatist talk about truths in the plural, about their utility and satisfactoriness, about the success with which they `work,' etc., suggests to the typical intellectualist mind a sort of coarse lame second-rate makeshift article of truth. Such truths are not real truth. Such tests are merely subjective. As against this, objective truth must be something non-utilitarian, haughty, refined, remote, august, exalted. It must be an absolute correspondence of our thoughts with an equally absolute reality. It must be what we \emph{ought} to think, unconditionally. The conditioned ways in which we \emph{do} think are so much irrelevance and matter for psychology. Down with psychology, up with logic, in all this question!

\wjparagraph
See the exquisite contrast of the types of mind! The pragmatist clings to facts and concreteness, observes truth at its work in particular cases, and generalizes. Truth, for him, becomes a class-name for all sorts of definite working-values in experience. For the rationalist it remains a pure abstraction, to the bare name of which we must defer. When the pragmatist undertakes to show in detail just \emph{why} we must defer, the rationalist is unable to recognize the concretes from which his own abstraction is taken. He accuses us of \emph{denying} truth; whereas we have only sought to trace exactly why people follow it and always ought to follow it. Your typical ultra-abstractionist fairly shudders at concreteness: other things equal, he positively prefers the pale and spectral. If the two universes were offered, he would always choose the skinny outline rather than the rich thicket of reality. It is so much purer, clearer, nobler.

\wjparagraph
I hope that as these lectures go on, the concreteness and closeness to facts of the pragmatism which they advocate may be what approves itself to you as its most satisfactory peculiarity. It only follows here the example of the sister-sciences, interpreting the unobserved by the observed. It brings old and new harmoniously together. It converts the absolutely empty notion of a static relation of `correspondence' (what that may mean we must ask later) between our minds and reality, into that of a rich and active commerce (that anyone may follow in detail and understand) between particular thoughts of ours, and the great universe of other experiences in which they play their parts and have their uses.

\wjparagraph
But enough of this at present? The justification of what I say must be postponed. I wish now to add a word in further explanation of the claim I made at our last meeting, that pragmatism may be a happy harmonizer of empiricist ways of thinking, with the more religious demands of human beings.

\wjparagraph
Men who are strongly of the fact-loving temperament, you may remember me to have said, are liable to be kept at a distance by the small sympathy with facts which that philosophy from the present-day fashion of idealism offers them. It is far too intellectualistic. Old fashioned theism was bad enough, with its notion of God as an exalted monarch, made up of a lot of unintelligible or preposterous `attributes'; but, so long as it held strongly by the argument from design, it kept some touch with concrete realities. Since, however, Darwinism has once for all displaced design from the minds of the `scientific,' theism has lost that foothold; and some kind of an immanent or pantheistic deity working \emph{in} things rather than above them is, if any, the kind recommended to our contemporary imagination. Aspirants to a philosophic religion turn, as a rule, more hopefully nowadays towards idealistic pantheism than towards the older dualistic theism, in spite of the fact that the latter still counts able defenders.

\wjparagraph
But, as I said in my first lecture, the brand of pantheism offered is hard for them to assimilate if they are lovers of facts, or empirically minded. It is the absolutistic brand, spurning the dust and reared upon pure logic. It keeps no connexion whatever with concreteness. Affirming the Absolute Mind, which is its substitute for God, to be the rational presupposition of all particulars of fact, whatever they may be, it remains supremely indifferent to what the particular facts in our world actually are. Be they what they may, the Absolute will father them. Like the sick lion in Esop's fable, all footprints lead into his den, but \emph{nulla vestigia retrorsum}.\footnote{[Latin for: no trace returning.]} You cannot redescend into the world of particulars by the Absolute's aid, or deduce any necessary consequences of detail important for your life from your idea of his nature. He gives you indeed the assurance that all is well with Him, and for his eternal way of thinking; but thereupon he leaves you to be finitely saved by your own temporal devices.

\wjparagraph
Far be it from me to deny the majesty of this conception, or its capacity to yield religious comfort to a most respectable class of minds. But from the human point of view, no one can pretend that it doesn't suffer from the faults of remoteness and abstractness. It is eminently a product of what I have ventured to call the rationalistic temper. It disdains empiricism's needs. It substitutes a pallid outline for the real world's richness. It is dapper; it is noble in the bad sense, in the sense in which to be noble is to be inapt for humble service. In this real world of sweat and dirt, it seems to me that when a view of things is `noble,' that ought to count as a presumption against its truth, and as a philosophic disqualification. The prince of darkness may be a gentleman, as we are told he is, but whatever the God of earth and heaven is, he can surely be no gentleman. His menial services are needed in the dust of our human trials, even more than his dignity is needed in the empyrean.

\wjparagraph
Now pragmatism, devoted tho she be to facts, has no such materialistic bias as ordinary empiricism labors under. Moreover, she has no objection whatever to the realizing of abstractions, so long as you get about among particulars with their aid and they actually carry you somewhere. Interested in no conclusions but those which our minds and our experiences work out together, she has no a priori prejudices against theology. \emph{If theological ideas prove to have a value for concrete life, they will be true, for pragmatism, in the sense of being good for so much. For how much more they are true, will depend entirely on their relations to the other truths that also have to be acknowledged.}

\wjparagraph
What I said just now about the Absolute of transcendental idealism is a case in point. First, I called it majestic and said it yielded religious comfort to a class of minds, and then I accused it of remoteness and sterility. But so far as it affords such comfort, it surely is not sterile; it has that amount of value; it performs a concrete function. As a good pragmatist, I myself ought to call the Absolute true `in so far forth,' then; and I unhesitatingly now do so.

\wjparagraph
But what does \emph{true in so far forth} mean in this case? To answer, we need only apply the pragmatic method. What do believers in the Absolute mean by saying that their belief affords them comfort? They mean that since in the Absolute finite evil is `overruled' already, we may, therefore, whenever we wish, treat the temporal as if it were potentially the eternal, be sure that we can trust its outcome, and, without sin, dismiss our fear and drop the worry of our finite responsibility. In short, they mean that we have a right ever and anon to take a moral holiday, to let the world wag in its own way, feeling that its issues are in better hands than ours and are none of our business.

\wjparagraph
The universe is a system of which the individual members may relax their anxieties occasionally, in which the don't-care mood is also right for men, and moral holidays in order--- that, if I mistake not, is part, at least, of what the Absolute is `known-as,' that is the great difference in our particular experiences which his being true makes for us, that is part of his cash-value when he is pragmatically interpreted. Farther than that the ordinary lay-reader in philosophy who thinks favorably of absolute idealism does not venture to sharpen his conceptions. He can use the Absolute for so much, and so much is very precious. He is pained at hearing you speak incredulously of the Absolute, therefore, and disregards your criticisms because they deal with aspects of the conception that he fails to follow.

\wjparagraph
If the Absolute means this, and means no more than this, who can possibly deny the truth of it? To deny it would be to insist that men should never relax, and that holidays are never in order. I am well aware how odd it must seem to some of you to hear me say that an idea is `true' so long as to believe it is profitable to our lives. That it is \emph{good}, for as much as it profits, you will gladly admit. If what we do by its aid is good, you will allow the idea itself to be good in so far forth, for we are the better for possessing it. But is it not a strange misuse of the word `truth,' you will say, to call ideas also `true' for this reason?

\wjparagraph
To answer this difficulty fully is impossible at this stage of my account. You touch here upon the very central point of Messrs. Schiller's, Dewey's and my own doctrine of truth, which I cannot discuss with detail until my sixth lecture. Let me now say only this, that truth is \emph{one species of good}, and not, as is usually supposed, a category distinct from good, and co-ordinate with it. \emph{The true is the name of whatever proves itself to be good in the way of belief, and good, too, for definite, assignable reasons.} Surely you must admit this, that if there were \emph{no} good for life in true ideas, or if the knowledge of them were positively disadvantageous and false ideas the only useful ones, then the current notion that truth is divine and precious, and its pursuit a duty, could never have grown up or become a dogma. In a world like that, our duty would be to \emph{shun} truth, rather. But in this world, just as certain foods are not only agreeable to our taste, but good for our teeth, our stomach and our tissues; so certain ideas are not only agreeable to think about, or agreeable as supporting other ideas that we are fond of, but they are also helpful in life's practical struggles. If there be any life that it is really better we should lead, and if there be any idea which, if believed in, would help us to lead that life, then it would be really \emph{better for us} to believe in that idea, \emph{unless, indeed, belief in it incidentally clashed with other greater vital benefits.}

\wjparagraph
`What would be better for us to believe'! This sounds very like a definition of truth. It comes very near to saying `what we \emph{ought} to believe': and in \emph{that} definition none of you would find any oddity. Ought we ever not to believe what it is \emph{better for us} to believe? And can we then keep the notion of what is better for us, and what is true for us, permanently apart?

\wjparagraph
Pragmatism says no, and I fully agree with her. Probably you also agree, so far as the abstract statement goes, but with a suspicion that if we practically did believe everything that made for good in our own personal lives, we should be found indulging all kinds of fancies about this world's affairs, and all kinds of sentimental superstitions about a world hereafter. Your suspicion here is undoubtedly well founded, and it is evident that something happens when you pass from the abstract to the concrete, that complicates the situation.

\wjparagraph
I said just now that what is better for us to believe is true \emph{unless the belief incidentally clashes with some other vital benefit}. Now in real life what vital benefits is any particular belief of ours most liable to clash with? What indeed except the vital benefits yielded by \emph{other beliefs} when these prove incompatible with the first ones? In other words, the greatest enemy of any one of our truths may be the rest of our truths. Truths have once for all this desperate instinct of self-preservation and of desire to extinguish whatever contradicts them. My belief in the Absolute, based on the good it does me, must run the gauntlet of all my other beliefs. Grant that it may be true in giving me a moral holiday. Nevertheless, as I conceive it,--- and let me speak now confidentially, as it were, and merely in my own private person,--- it clashes with other truths of mine whose benefits I hate to give up on its account. It happens to be associated with a kind of logic of which I am the enemy, I find that it entangles me in metaphysical paradoxes that are inacceptable, etc., etc.. But as I have enough trouble in life already without adding the trouble of carrying these intellectual inconsistencies, I personally just give up the Absolute. I just \emph{take} my moral holidays; or else as a professional philosopher, I try to justify them by some other principle.

\wjparagraph
If I could restrict my notion of the Absolute to its bare holiday-giving value, it wouldn't clash with my other truths. But we cannot easily thus restrict our hypotheses. They carry supernumerary features, and these it is that clash so. My disbelief in the Absolute means then disbelief in those other supernumerary features, for I fully believe in the legitimacy of taking moral holidays.

\wjparagraph
You see by this what I meant when I called pragmatism a mediator and reconciler and said, borrowing the word from Papini, that she unstiffens our theories. She has in fact no prejudices whatever, no obstructive dogmas, no rigid canons of what shall count as proof. She is completely genial. She will entertain any hypothesis, she will consider any evidence. It follows that in the religious field she is at a great advantage both over positivistic empiricism, with its anti-theological bias, and over religious rationalism, with its exclusive interest in the remote, the noble, the simple, and the abstract in the way of conception.

\wjparagraph
In short, she widens the field of search for God. Rationalism sticks to logic and the empyrean. Empiricism sticks to the external senses. Pragmatism is willing to take anything, to follow either logic or the senses, and to count the humblest and most personal experiences. She will count mystical experiences if they have practical consequences. She will take a God who lives in the very dirt of private fact--- if that should seem a likely place to find him.

\wjparagraph
Her only test of probable truth is what works best in the way of leading us, what fits every part of life best and combines with the collectivity of experience's demands, nothing being omitted. If theological ideas should do this, if the notion of God, in particular, should prove to do it, how could pragmatism possibly deny God's existence? She could see no meaning in treating as `not true' a notion that was pragmatically so successful. What other kind of truth could there be, for her, than all this agreement with concrete reality?

\wjparagraph
In my last lecture I shall return again to the relations of pragmatism with religion. But you see already how democratic she is. Her manners are as various and flexible, her resources as rich and endless, and her conclusions as friendly as those of mother nature.


\section*{Lecture III: Some Metaphysical Problems Pragmatically Considered}
\setcounter{wjparagraph}{0}

\wjparagraph
I am now to make the pragmatic method more familiar by giving you some illustrations of its application to particular problems. I will begin with what is driest, and the first thing I shall take will be the problem of Substance. Everyone uses the old distinction between substance and attribute, enshrined as it is in the very structure of human language, in the difference between grammatical subject and predicate. Here is a bit of blackboard crayon. Its modes, attributes, properties, accidents, or affections,--- use which term you will,--- are whiteness, friability, cylindrical shape, insolubility in water, etc., etc. But the bearer of these attributes is so much chalk, which thereupon is called the substance in which they inhere. So the attributes of this desk inhere in the substance `wood,' those of my coat in the substance `wool,' and so forth. Chalk, wood and wool, show again, in spite of their differences, common properties, and in so far forth they are themselves counted as modes of a still more primal substance, matter, the attributes of which are space occupancy and impenetrability. Similarly our thoughts and feelings are affections or properties of our several souls, which are substances, but again not wholly in their own right, for they are modes of the still deeper substance `spirit.'

\wjparagraph
Now it was very early seen that all we know of the chalk is the whiteness, friability, etc., all \emph{we know} of the wood is the combustibility and fibrous structure. A group of attributes is what each substance here is known-as, they form its sole cash-value for our actual experience. The substance is in every case revealed through \emph{them}; if we were cut off from \emph{them} we should never suspect its existence; and if God should keep sending them to us in an unchanged order, miraculously annihilating at a certain moment the substance that supported them, we never could detect the moment, for our experiences themselves would be unaltered. Nominalists accordingly adopt the opinion that substance is a spurious idea due to our inveterate human trick of turning names into things. Phenomena come in groups--- the chalk-group, the wood-group, etc.--- and each group gets its name. The name we then treat as in a way supporting the group of phenomena. The low thermometer to-day, for instance, is supposed to come from something called the `climate.' Climate is really only the name for a certain group of days, but it is treated as if it lay \emph{behind} the day, and in general we place the name, as if it were a being, behind the facts it is the name of. But the phenomenal properties of things, nominalists say, surely do not really inhere in names, and if not in names then they do not inhere in anything. They \emph{ad}here, or \emph{co}here, rather, \emph{with each other}, and the notion of a substance inaccessible to us, which we think accounts for such cohesion by supporting it, as cement might support pieces of mosaic, must be abandoned. The fact of the bare cohesion itself is all that the notion of the substance signifies. Behind that fact is nothing.

\wjparagraph
Scholasticism has taken the notion of substance from common sense and made it very technical and articulate. Few things would seem to have fewer pragmatic consequences for us than substances, cut off as we are from every contact with them. Yet in one case scholasticism has proved the importance of the substance-idea by treating it pragmatically. I refer to certain disputes about the mystery of the Eucharist. Substance here would appear to have momentous pragmatic value. Since the accidents of the wafer don't change in the Lord's supper, and yet it has become the very body of Christ, it must be that the change is in the substance solely. The bread-substance must have been withdrawn, and the divine substance substituted miraculously without altering the immediate sensible properties. But tho these don't alter, a tremendous difference has been made, no less a one than this, that we who take the sacrament, now feed upon the very substance of divinity. The substance-notion breaks into life, then, with tremendous effect, if once you allow that substances can separate from their accidents, and exchange these latter.

\wjparagraph
This is the only pragmatic application of the substance-idea with which I am acquainted; and it is obvious that it will only be treated seriously by those who already believe in the `real presence' on independent grounds.

\wjparagraph
\emph{Material substance} was criticized by Berkeley with such telling effect that his name has reverberated through all subsequent philosophy. Berkeley's treatment of the notion of matter is so well known as to need hardly more than a mention. So far from denying the external world which we know, Berkeley corroborated it. It was the scholastic notion of a material substance unapproachable by us, \emph{behind} the external world, deeper and more real than it, and needed to support it, which Berkeley maintained to be the most effective of all reducers of the external world to unreality. Abolish that substance, he said, believe that God, whom you can understand and approach, sends you the sensible world directly, and you confirm the latter and back it up by his divine authority. Berkeley's criticism of `matter' was consequently absolutely pragmatistic. Matter is known as our sensations of colour, figure, hardness and the like. They are the cash-value of the term. The difference matter makes to us by truly being is that we then get such sensations; by not being, is that we lack them. These sensations then are its sole meaning. Berkeley doesn't deny matter, then; he simply tells us what it consists of. It is a true name for just so much in the way of sensations.

\wjparagraph
Locke, and later Hume, applied a similar pragmatic criticism to the notion of \emph{spiritual substance}. I will only mention Locke's treatment of our `personal identity.' He immediately reduces this notion to its pragmatic value in terms of experience. It means, he says, so much `consciousness,' namely the fact that at one moment of life we remember other moments, and feel them all as parts of one and the same personal history. Rationalism had explained this practical continuity in our life by the unity of our soul-substance. But Locke says: suppose that God should take away the consciousness, should \emph{we} be any the better for having still the soul-principle? Suppose he annexed the same consciousness to different souls, should \emph{we}, as we realize \emph{ourselves}, be any the worse for that fact? In Locke's day the soul was chiefly a thing to be rewarded or punished. See how Locke, discussing it from this point of view, keeps the question pragmatic:

\wjparagraph
Suppose, he says, one to think himself to be the same soul that once was Nestor or Thersites. Can he think their actions his own any more than the actions of any other man that ever existed? But let him once find himself \emph{conscious} of any of the actions of Nestor, he then finds himself the same person with Nestor. ... In this personal identity is founded all the right and justice of reward and punishment. It may be reasonable to think, no one shall be made to answer for what he knows nothing of, but shall receive his doom, his consciousness accusing or excusing. Supposing a man punished now for what he had done in another life, whereof he could be made to have no consciousness at all, what difference is there between that punishment and being created miserable?

\wjparagraph
Our personal identity, then, consists, for Locke, solely in pragmatically definable particulars. Whether, apart from these verifiable facts, it also inheres in a spiritual principle, is a merely curious speculation. Locke, compromiser that he was, passively tolerated the belief in a substantial soul behind our consciousness. But his successor Hume, and most empirical psychologists after him, have denied the soul, save as the name for verifiable cohesions in our inner life. They redescend into the stream of experience with it, and cash it into so much small-change value in the way of `ideas' and their peculiar connexions with each other. As I said of Berkeley's matter, the soul is good or `true' for just \emph{so much}, but no more.

\wjparagraph
The mention of material substance naturally suggests the doctrine of `materialism,' but philosophical materialism is not necessarily knit up with belief in `matter,' as a metaphysical principle. One may deny matter in that sense, as strongly as Berkeley did, one may be a phenomenalist like Huxley, and yet one may still be a materialist in the wider sense, of explaining higher phenomena by lower ones, and leaving the destinies of the world at the mercy of its blinder parts and forces. It is in this wider sense of the word that materialism is opposed to spiritualism or theism. The laws of physical nature are what run things, materialism says. The highest productions of human genius might be ciphered by one who had complete acquaintance with the facts, out of their physiological conditions, regardless whether nature be there only for our minds, as idealists contend, or not. Our minds in any case would have to record the kind of nature it is, and write it down as operating through blind laws of physics. This is the complexion of present day materialism, which may better be called naturalism. Over against it stands `theism,' or what in a wide sense may be termed `spiritualism.' Spiritualism says that mind not only witnesses and records things, but also runs and operates them: the world being thus guided, not by its lower, but by its higher element.

\wjparagraph
Treated as it often is, this question becomes little more than a conflict between aesthetic preferences. Matter is gross, coarse, crass, muddy; spirit is pure, elevated, noble; and since it is more consonant with the dignity of the universe to give the primacy in it to what appears superior, spirit must be affirmed as the ruling principle. To treat abstract principles as finalities, before which our intellects may come to rest in a state of admiring contemplation, is the great rationalist failing. Spiritualism, as often held, may be simply a state of admiration for one kind, and of dislike for another kind, of abstraction. I remember a worthy spiritualist professor who always referred to materialism as the `mud-philosophy,' and deemed it thereby refuted.

\wjparagraph
To such spiritualism as this there is an easy answer, and Mr. Spencer makes it effectively. In some well-written pages at the end of the first volume of his Psychology he shows us that a `matter' so infinitely subtile, and performing motions as inconceivably quick and fine as those which modern science postulates in her explanations, has no trace of grossness left. He shows that the conception of spirit, as we mortals hitherto have framed it, is itself too gross to cover the exquisite tenuity of nature's facts. Both terms, he says, are but symbols, pointing to that one unknowable reality in which their oppositions cease.

\wjparagraph
To an abstract objection an abstract rejoinder suffices; and so far as one's opposition to materialism springs from one's disdain of matter as something `crass,' Mr. Spencer cuts the ground from under one. Matter is indeed infinitely and incredibly refined. To anyone who has ever looked on the face of a dead child or parent the mere fact that matter \emph{could} have taken for a time that precious form, ought to make matter sacred ever after. It makes no difference what the \emph{principle} of life may be, material or immaterial, matter at any rate co-operates, lends itself to all life's purposes. That beloved incarnation was among matter's possibilities.

\wjparagraph
But now, instead of resting in principles after this stagnant intellectualist fashion, let us apply the pragmatic method to the question. What do we \emph{mean} by matter? What practical difference can it make \emph{now} that the world should be run by matter or by spirit? I think we find that the problem takes with this a rather different character.

\wjparagraph
And first of all I call your attention to a curious fact. It makes not a single jot of difference so far as the \emph{past} of the world goes, whether we deem it to have been the work of matter or whether we think a divine spirit was its author.

\wjparagraph
Imagine, in fact, the entire contents of the world to be once for all irrevocably given. Imagine it to end this very moment, and to have no future; and then let a theist and a materialist apply their rival explanations to its history. The theist shows how a God made it; the materialist shows, and we will suppose with equal success, how it resulted from blind physical forces. Then let the pragmatist be asked to choose between their theories. How can he apply his test if the world is already completed? Concepts for him are things to come back into experience with, things to make us look for differences. But by hypothesis there is to be no more experience and no possible differences can now be looked for. Both theories have shown all their consequences and, by the hypothesis we are adopting, these are identical. The pragmatist must consequently say that the two theories, in spite of their different-sounding names, mean exactly the same thing, and that the dispute is purely verbal. [I am supposing, of course, that the theories \emph{have} been equally successful in their explanations of what is.]

\wjparagraph
For just consider the case sincerely, and say what would be the \emph{worth} of a God if he \emph{were} there, with his work accomplished and his world run down. He would be worth no more than just that world was worth. To that amount of result, with its mixed merits and defects, his creative power could attain, but go no farther. And since there is to be no future; since the whole value and meaning of the world has been already paid in and actualized in the feelings that went with it in the passing, and now go with it in the ending; since it draws no supplemental significance (such as our real world draws) from its function of preparing something yet to come; why then, by it we take God's measure, as it were. He is the Being who could once for all do \emph{that}; and for that much we are thankful to him, but for nothing more. But now, on the contrary hypothesis, namely, that the bits of matter following their laws could make that world and do no less, should we not be just as thankful to them? Wherein should we suffer loss, then, if we dropped God as an hypothesis and made the matter alone responsible? Where would any special deadness, or crassness, come in? And how, experience being what is once for all, would God's presence in it make it any more living or richer?

\wjparagraph
Candidly, it is impossible to give any answer to this question. The actually experienced world is supposed to be the same in its details on either hypothesis, ``the same, for our praise or blame,'' as Browning says. It stands there indefeasibly: a gift which can't be taken back. Calling matter the cause of it retracts no single one of the items that have made it up, nor does calling God the cause augment them. They are the God or the atoms, respectively, of just that and no other world. The God, if there, has been doing just what atoms could do--- appearing in the character of atoms, so to speak--- and earning such gratitude as is due to atoms, and no more. If his presence lends no different turn or issue to the performance, it surely can lend it no increase of dignity. Nor would indignity come to it were he absent, and did the atoms remain the only actors on the stage. When a play is once over, and the curtain down, you really make it no better by claiming an illustrious genius for its author, just as you make it no worse by calling him a common hack.

\wjparagraph
Thus if no future detail of experience or conduct is to be deduced from our hypothesis, the debate between materialism and theism becomes quite idle and insignificant. Matter and God in that event mean exactly the same thing--- the power, namely, neither more nor less, that could make just this completed world--- and the wise man is he who in such a case would turn his back on such a supererogatory discussion. Accordingly, most men instinctively, and positivists and scientists deliberately, do turn their backs on philosophical disputes from which nothing in the line of definite future consequences can be seen to follow. The verbal and empty character of philosophy is surely a reproach with which we are, but too familiar. If pragmatism be true, it is a perfectly sound reproach unless the theories under fire can be shown to have alternative practical outcomes, however delicate and distant these may be. The common man and the scientist say they discover no such outcomes, and if the metaphysician can discern none either, the others certainly are in the right of it, as against him. His science is then but pompous trifling; and the endowment of a professorship for such a being would be silly.

\wjparagraph
Accordingly, in every genuine metaphysical debate some practical issue, however conjectural and remote, is involved. To realize this, revert with me to our question, and place yourselves this time in the world we live in, in the world that \emph{has} a future, that is yet uncompleted whilst we speak. In this unfinished world the alternative of `materialism or theism?' is intensely practical; and it is worth while for us to spend some minutes of our hour in seeing that it is so.

\wjparagraph
How, indeed, does the program differ for us, according as we consider that the facts of experience up to date are purposeless configurations of blind atoms moving according to eternal laws, or that on the other hand they are due to the providence of God? As far as the past facts go, indeed there is no difference. Those facts are in, are bagged, are captured; and the good that's in them is gained, be the atoms or be the God their cause. There are accordingly many materialists about us to-day who, ignoring altogether the future and practical aspects of the question, seek to eliminate the odium attaching to the word materialism, and even to eliminate the word itself, by showing that, if matter could give birth to all these gains, why then matter, functionally considered, is just as divine an entity as God, in fact coalesces with God, is what you mean by God. Cease, these persons advise us, to use either of these terms, with their outgrown opposition. Use a term free of the clerical connotations, on the one hand; of the suggestion of gross-ness, coarseness, ignobility, on the other. Talk of the primal mystery, of the unknowable energy, of the one and only power, instead of saying either God or matter. This is the course to which Mr. Spencer urges us; and if philosophy were purely retrospective, he would thereby proclaim himself an excellent pragmatist.

\wjparagraph
But philosophy is prospective also, and, after finding what the world has been and done and yielded, still asks the further question `what does the world \emph{promise}?' Give us a matter that promises \emph{success}, that is bound by its laws to lead our world ever nearer to perfection, and any rational man will worship that matter as readily as Mr. Spencer worships his own so-called unknowable power. It not only has made for righteousness up to date, but it will make for righteousness forever; and that is all we need. Doing practically all that a God can do, it is equivalent to God, its function is a God's function, and is exerted in a world in which a God would now be superfluous; from such a world a God could never lawfully be missed. `Cosmic emotion' would here be the right name for religion.

\wjparagraph
But is the matter by which Mr. Spencer's process of cosmic evolution is carried on any such principle of never-ending perfection as this? Indeed it is not, for the future end of every cosmically evolved thing or system of things is foretold by science to be death and tragedy; and Mr. Spencer, in confining himself to the aesthetic and ignoring the practical side of the controversy, has really contributed nothing serious to its relief. But apply now our principle of practical results, and see what a vital significance the question of materialism or theism immediately acquires.

\wjparagraph
Theism and materialism, so indifferent when taken retrospectively, point, when we take them prospectively, to wholly different outlooks of experience. For, according to the theory of mechanical evolution, the laws of redistribution of matter and motion, tho they are certainly to thank for all the good hours which our organisms have ever yielded us and for all the ideals which our minds now frame, are yet fatally certain to undo their work again, and to redissolve everything that they have once evolved. You all know the picture of the last state of the universe which evolutionary science foresees. I cannot state it better than in Mr. Balfour's words: ``The energies of our system will decay, the glory of the sun will be dimmed, and the earth, tideless and inert, will no longer tolerate the race which has for a moment disturbed its solitude. Man will go down into the pit, and all his thoughts will perish. The uneasy, consciousness which in this obscure corner has for a brief space broken the contented silence of the universe, will be at rest. Matter will know itself no longer. `Imperishable monuments' and `immortal deeds,' death itself, and love stronger than death, will be as though they had never been. Nor will anything that is, be better or be worse for all that the labour, genius, devotion, and suffering of man have striven through countless generations to effect.''\footnote{The Foundations of Belief, p. 30.}

\wjparagraph
That is the sting of it, that in the vast driftings of the cosmic weather, tho many a jeweled shore appears, and many an enchanted cloud-bank floats away, long lingering ere it be dissolved--- even as our world now lingers, for our joy-yet when these transient products are gone, nothing, absolutely \emph{nothing} remains, of represent those particular qualities, those elements of preciousness which they may have enshrined. Dead and gone are they, gone utterly from the very sphere and room of being. Without an echo; without a memory; without an influence on aught that may come after, to make it care for similar ideals. This utter final wreck and tragedy is of the essence of scientific materialism as at present understood. The lower and not the higher forces are the eternal forces, or the last surviving forces within the only cycle of evolution which we can definitely see. Mr. Spencer believes this as much as anyone; so why should he argue with us as if we were making silly aesthetic objections to the `grossness' of `matter and motion,' the principles of his philosophy, when what really dismays us is the disconsolateness of its ulterior practical results?

\wjparagraph
No the true objection to materialism is not positive but negative. It would be farcical at this day to make complaint of it for what it \emph{is} for `grossness.' Grossness is what grossness \emph{does}--- we now know \emph{that}. We make complaint of it, on the contrary, for what it is \emph{not}--- not a permanent warrant for our more ideal interests, not a fulfiller of our remotest hopes.

\wjparagraph
The notion of God, on the other hand, however inferior it may be in clearness to those mathematical notions so current in mechanical philosophy, has at least this practical superiority over them, that it guarantees an ideal order that shall be permanently preserved. A world with a God in it to say the last word, may indeed burn up or freeze, but we then think of him as still mindful of the old ideals and sure to bring them elsewhere to fruition; so that, where he is, tragedy is only provisional and partial, and shipwreck and dissolution not the absolutely final things. This need of an eternal moral order is one of the deepest needs of our breast. And those poets, like Dante and Wordsworth, who live on the conviction of such an order, owe to that fact the extraordinary tonic and consoling power of their verse. Here then, in these different emotional and practical appeals, in these adjustments of our concrete attitudes of hope and expectation, and all the delicate consequences which their differences entail, lie the real meanings of materialism and spiritualism--- not in hair-splitting abstractions about matter's inner essence, or about the metaphysical attributes of God. Materialism means simply the denial that the moral order is eternal, and the cutting off of ultimate hopes; spiritualism means the affirmation of an eternal moral order and the letting loose of hope. Surely here is an issue genuine enough, for anyone who feels it; and, as long as men are men, it will yield matter for a serious philosophic debate.

\wjparagraph
But possibly some of you may still rally to their defence. Even whilst admitting that spiritualism and materialism make different prophecies of the world's future, you may yourselves pooh-pooh the difference as something so infinitely remote as to mean nothing for a sane mind. The essence of a sane mind, you may say, is to take shorter views, and to feel no concern about such chimaeras as the latter end of the world. Well, I can only say that if you say this, you do injustice to human nature. Religious melancholy is not disposed of by a simple flourish of the word insanity. The absolute things, the last things, the overlapping things, are the truly philosophic concerns; all superior minds feel seriously about them, and the mind with the shortest views is simply the mind of the more shallow man.

\wjparagraph
The issues of fact at stake in the debate are of course vaguely enough conceived by us at present. But spiritualistic faith in all its forms deals with a world of \emph{promise}, while materialism's sun sets in a sea of disappointment. Remember what I said of the Absolute: it grants us moral holidays. Any religious view does this. It not only incites our more strenuous moments, but it also takes our joyous, careless, trustful moments, and it justifies them. It paints the grounds of justification vaguely enough, to be sure. The exact features of the saving future facts that our belief in God insures, will have to be ciphered out by the interminable methods of science: we can \emph{study} our God only by studying his Creation. But we can \emph{enjoy} our God, if we have one, in advance of all that labor. I myself believe that the evidence for God lies primarily in inner personal experiences. When they have once given you your God, his name means at least the benefit of the holiday. You remember what I said yesterday about the way in which truths clash and try to `down' each other. The truth of `God' has to run the gauntlet of all our other truths. It is on trial by them and they on trial by it. Our \emph{final} opinion about God can be settled only after all the truths have straightened themselves out together. Let us hope that they shall find a modus vivendi!

\wjparagraph
Let me pass to a very cognate philosophic problem, the \emph{question of design in nature}. God's existence has from time immemorial been held to be proved by certain natural facts. Many facts appear as if expressly designed in view of one another. Thus the woodpecker's bill, tongue, feet, tail, etc., fit him wondrously for a world of trees with grubs hid in their bark to feed upon. The parts of our eye fit the laws of light to perfection, leading its rays to a sharp picture on our retina. Such mutual fitting of things diverse in origin argued design, it was held; and the designer was always treated as a man-loving deity.

\wjparagraph
The first step in these arguments was to prove that the design existed. Nature was ransacked for results obtained through separate things being co-adapted. Our eyes, for instance, originate in intra-uterine darkness, and the light originates in the sun, yet see how they fit each other. They are evidently made \emph{for} each other. Vision is the end designed, light and eyes the separate means devised for its attainment.

\wjparagraph
It is strange, considering how unanimously our ancestors felt the force of this argument, to see how little it counts for since the triumph of the Darwinian theory. Darwin opened our minds to the power of chance-happenings to bring forth `fit' results if only they have time to add themselves together. He showed the enormous waste of nature in producing results that get destroyed because of their unfitness. He also emphasized the number of adaptations which, if designed, would argue an evil rather than a good designer. Here all depends upon the point of view. To the grub under the bark the exquisite fitness of the woodpecker's organism to extract him would certainly argue a diabolical designer.

\wjparagraph
Theologians have by this time stretched their minds so as to embrace the Darwinian facts, and yet to interpret them as still showing divine purpose. It used to be a question of purpose \emph{against} mechanism, of one \emph{or} the other. It was as if one should say ``My shoes are evidently designed to fit my feet, hence it is impossible that they should have been produced by machinery.'' We know that they are both: they are made by a machinery itself designed to fit the feet with shoes. Theology need only stretch similarly the designs of God. As the aim of a football-team is not merely to get the ball to a certain goal (if that were so, they would simply get up on some dark night and place it there), but to get it there by a fixed \emph{machinery of conditions}--- the game's rules and the opposing players; so the aim of God is not merely, let us say, to make men and to save them, but rather to get this done through the sole agency of nature's vast machinery. Without nature's stupendous laws and counterforces, man's creation and perfection, we might suppose, would be too insipid achievements for God to have designed them.

\wjparagraph
This saves the form of the design-argument at the expense of its old easy human content. The designer is no longer the old man-like deity. His designs have grown so vast as to be incomprehensible to us humans. The \emph{what} of them so overwhelms us that to establish the mere \emph{that} of a designer for them becomes of very little consequence in comparison. We can with difficulty comprehend the character of a cosmic mind whose purposes are fully revealed by the strange mixture of goods and evils that we find in this actual world's particulars. Or rather we cannot by any possibility comprehend it. The mere word `design' by itself has, we see, no consequences and explains nothing. It is the barrenest of principles. The old question of \emph{whether} there is design is idle. The real question is \emph{what} is the world, whether or not it have a designer--- and that can be revealed only by the study of all nature's particulars.

\wjparagraph
Remember that no matter what nature may have produced or may be producing, the means must necessarily have been adequate, must have been \emph{fitted to that production}. The argument from fitness to design would consequently always apply, whatever were the product's character. The recent Mont-Pelee eruption, for example, required all previous history to produce that exact combination of ruined houses, human and animal corpses, sunken ships, volcanic ashes, etc., in just that one hideous configuration of positions. France had to be a nation and colonize Martinique. Our country had to exist and send our ships there. \emph{If} God aimed at just that result, the means by which the centuries bent their influences towards it, showed exquisite intelligence. And so of any state of things whatever, either in nature or in history, which we find actually realized. For the parts of things must always make \emph{some} definite resultant, be it chaotic or harmonious. When we look at what has actually come, the conditions must always appear perfectly designed to ensure it. We can always say, therefore, in any conceivable world, of any conceivable character, that the whole cosmic machinery \emph{may} have been designed to produce it.

\wjparagraph
Pragmatically, then, the abstract word `design' is a blank cartridge. It carries no consequences, it does no execution. What sort of design? and what sort of a designer? are the only serious questions, and the study of facts is the only way of getting even approximate answers. Meanwhile, pending the slow answer from facts, anyone who insists that there is a designer and who is sure he is a divine one, gets a certain pragmatic benefit from the term--- the same, in fact which we saw that the terms God, Spirit, or the Absolute, yield us `Design,' worthless tho it be as a mere rationalistic principle set above or behind things for our admiration, becomes, if our faith concretes it into something theistic, a term of \emph{promise}. Returning with it into experience, we gain a more confiding outlook on the future. If not a blind force but a seeing force runs things, we may reasonably expect better issues. This vague confidence in the future is the sole pragmatic meaning at present discernible in the terms design and designer. But if cosmic confidence is right not wrong, better not worse, that is a most important meaning. That much at least of possible `truth' the terms will then have in them.

\wjparagraph
Let me take up another well-worn controversy, \emph{the free-will problem}. Most persons who believe in what is called their free-will do so after the rationalistic fashion. It is a principle, a positive faculty or virtue added to man, by which his dignity is enigmatically augmented. He ought to believe it for this reason. Determinists, who deny it, who say that individual men originate nothing, but merely transmit to the future the whole push of the past cosmos of which they are so small an expression, diminish man. He is less admirable, stripped of this creative principle. I imagine that more than half of you share our instinctive belief in free-will, and that admiration of it as a principle of dignity has much to do with your fidelity.

\wjparagraph
But free-will has also been discussed pragmatically, and, strangely enough, the same pragmatic interpretation has been put upon it by both disputants. You know how large a part questions of \emph{accountability} have played in ethical controversy. To hear some persons, one would suppose that all that ethics aims at is a code of merits and demerits. Thus does the old legal and theological leaven, the interest in crime and sin and punishment abide with us. `Who's to blame? whom can we punish? whom will God punish?'--- these preoccupations hang like a bad dream over man's religious history.

\wjparagraph
So both free-will and determinism have been inveighed against and called absurd, because each, in the eyes of its enemies, has seemed to prevent the `imputability' of good or bad deeds to their authors. Queer antinomy this! Free-will means novelty, the grafting on to the past of something not involved therein. If our acts were predetermined, if we merely transmitted the push of the whole past, the free-willists say, how could we be praised or blamed for anything? We should be `agents' only, not `principals,' and where then would be our precious imputability and responsibility?

\wjparagraph
But where would it be if we \emph{has} free-will? rejoin the determinists. If a `free' act be a sheer novelty, that comes not \emph{from} me, the previous me, but ex nihilo, and simply tacks itself on to me, how can I, the previous I, be responsible? How can I have any permanent \emph{character} that will stand still long enough for praise or blame to be awarded? The chaplet of my days tumbles into a cast of disconnected beads as soon as the thread of inner necessity is drawn out by the preposterous indeterminist doctrine. Messrs. Fullerton and McTaggart have recently laid about them doughtily with this argument.

\wjparagraph
It may be good ad hominem, but otherwise it is pitiful. For I ask you, quite apart from other reasons, whether any man, woman or child, with a sense for realities, ought not to be ashamed to plead such principles as either dignity or imputability. Instinct and utility between them can safely be trusted to carry on the social business of punishment and praise. If a man does good acts we shall praise him, if he does bad acts we shall punish him--- anyhow, and quite apart from theories as to whether the acts result from what was previous in him or are novelties in a strict sense. To make our human ethics revolve about the question of `merit' is a piteous unreality--- God alone can know our merits, if we have any. The real ground for supposing free-will is indeed pragmatic, but it has nothing to do with this contemptible right to punish which had made such a noise in past discussions of the subject.

\wjparagraph
Free-will pragmatically means \emph{novelties in the world}, the right to expect that in its deepest elements as well as in its surface phenomena, the future may not identically repeat and imitate the past. That imitation en masse is there, who can deny? The general `uniformity of nature' is presupposed by every lesser law. But nature may be only approximately uniform; and persons in whom knowledge of the world's past has bred pessimism (or doubts as to the world's good character, which become certainties if that character be supposed eternally fixed) may naturally welcome free-will as a \emph{melioristic} doctrine. It holds up improvement as at least possible; whereas determinism assures us that our whole notion of possibility is born of human ignorance, and that necessity and impossibility between them rule the destinies of the world.

\wjparagraph
Free-will is thus a general cosmological theory of \emph{promise}, just like the Absolute, God, Spirit or Design. Taken abstractly, no one of these terms has any inner content, none of them gives us any picture, and no one of them would retain the least pragmatic value in a world whose character was obviously perfect from the start. Elation at mere existence, pure cosmic emotion and delight, would, it seems to me, quench all interest in those speculations, if the world were nothing but a lubberland of happiness already. Our interest in religious metaphysics arises in the fact that our empirical future feels to us unsafe, and needs some higher guarantee. If the past and present were purely good, who could wish that the future might possibly not resemble them? Who could desire free-will? Who would not say, with Huxley, ``let me be wound up every day like a watch, to go right fatally, and I ask no better freedom.'' `Freedom' in a world already perfect could only mean freedom to \emph{be worse}, and who could be so insane as to wish that? To be necessarily what it is, to be impossibly aught else, would put the last touch of perfection upon optimism's universe. Surely the only \emph{possibility} that one can rationally claim is the possibility that things may be \emph{better}. That possibility, I need hardly say, is one that, as the actual world goes, we have ample grounds for desiderating.

\wjparagraph
Free-will thus has no meaning unless it be a doctrine of \emph{relief}. As such, it takes its place with other religious doctrines. Between them, they build up the old wastes and repair the former desolations. Our spirit, shut within this courtyard of sense-experience, is always saying to the intellect upon the tower: `Watchman, tell us of the night, if it aught of promise bear,' and the intellect gives it then these terms of promise.

\wjparagraph
Other than this practical significance, the words God, free-will, design, etc., have none. Yet dark tho they be in themselves, or intellectualistically taken, when we bear them into life's thicket with us the darkness \emph{there} grows light about us. If you stop, in dealing with such words, with their definition, thinking that to be an intellectual finality, where are you? Stupidly staring at a pretentious sham! ``Deus est Ens, a se, extra et supra omne genus, necessarium, unum, infinite perfectum, simplex, immutabile, immensum, aeternum, intelligens,''\footnote{[Literally: God is a being, from itself, above and beyond all else, necessary, one, infinitely perfect, simple, immutable, immense, eternal, understanding.]}  etc.,--- wherein is such a definition really instructive? It means less, than nothing, in its pompous robe of adjectives. Pragmatism alone can read a positive meaning into it, and for that she turns her back upon the intellectualist point of view altogether. `God's in his heaven; all's right with the world!'--- \emph{that's} the heart of your theology, and for that you need no rationalist definitions.

\wjparagraph
Why shouldn't we all of us, rationalists as well as pragmatists, confess this? Pragmatism, so far from keeping her eyes bent on the immediate practical foreground, as she is accused of doing, dwells just as much upon the world's remotest perspectives.

\wjparagraph
See then how all these ultimate questions turn, as it were, up their hinges; and from looking backwards upon principles, upon an \emph{erkenntnisstheoretische Ich}, a God, a \emph{Kausalit\"atsprinzip}, a Design, a Free-will, taken in themselves, as something august and exalted above facts,--- see, I say, how pragmatism shifts the emphasis and looks forward into facts themselves. The really vital question for us all is, What is this world going to be? What is life eventually to make of itself? The centre of gravity of philosophy must therefore alter its place. The earth of things, long thrown into shadow by the glories of the upper ether, must resume its rights. To shift the emphasis in this way means that philosophic questions will fall to be treated by minds of a less abstractionist type than heretofore, minds more scientific and individualistic in their tone yet not irreligious either. It will be an alteration in `the seat of authority' that reminds one almost of the protestant reformation. And as, to papal minds, protestantism has often seemed a mere mess of anarchy and confusion, such, no doubt, will pragmatism often seem to ultra-rationalist minds in philosophy. It will seem so much sheer trash, philosophically. But life wags on, all the same, and compasses its ends, in protestant countries. I venture to think that philosophic protestantism will compass a not dissimilar prosperity. 

\end{document}