\documentclass[12pt]{article}
\usepackage{hyperref,setspace,graphicx,fancyhdr}
\usepackage[]{accessibility}
\usepackage[width=6in,height=9in]{geometry}
% changes font to TeX Gyre Schola (Century Schoolbook)
\usepackage{tgschola}\usepackage[T1]{fontenc}

\newcounter{authornote}[page]
\newcommand{\authornote}[1]{\renewcommand{\thefootnote}{\fnsymbol{footnote}}\stepcounter{authornote}\footnote[\value{authornote}]{#1}\renewcommand{\thefootnote}{\arabic{footnote}}}

\newcommand{\hbreak}{\par\noindent\begin{tabular*}{\linewidth}{c}\hline\hline\end{tabular*}\par}

\newcommand*{\authortitle}[1]{\centerline{\Huge\sc #1}}

\begin{document}
\authortitle{C.S. Peirce}

\bigskip

\paragraph{About Peirce:}
Charles Sanders Peirce (1839--1914) was born in Cambridge,
Massachusetts. (His last name rhymes with \emph{purse}, not with \emph{fierce}.) His father, Benjamin Peirce. was Professor of Mathematics at
Harvard University, one of the founders of the U.S. Coast and
Geodetic Survey, and one of the founders of the Smithsonian
Institution.

Peirce graduated from Harvard in 1859. He was then
employed by the U.S. Coast and Geodetic Survey, mainly surveying and
carrying out geodetic investigations. From 1879 through 1884, Peirce
also taught logic in the Department of Mathematics at Johns Hopkins University. This job ended suddenly for reasons that
apparently connected to Peirce cohabiting with his wife before marriage. Peirce's position with the the Coast and Geodetic Survey was
terminated in late 1891 because of a congressional funding decision. For the remainder of his life, Peirce eked out a living doing odd-jobs and consulting work--- mainly in chemical engineering and analysis. Peirce was often in dire financial straits; at times he managed to survive only through the charity of friends, including his old college chum William James.

Peirce's writings extend from about 1857 to near his
death, his published works run to about 12,000 printed pages, and his known unpublished manuscripts run to about 80,000 handwritten pages. The topics on which he wrote have an immense range: mathematics, logic, metaphysics, physical science, economics, the theory of signs\ldots

\paragraph{About this text:}
This document is prepared by \href{https://www.fecundity.com}{P.D. Magnus} from various sources. The biography is adapted from Robert Burch's contribution to the \href{http://plato.stanford.edu}{\emph{Stanford Encyclopedia of Philosophy}}.

\input{Peirce-from-some-consequences.tex}
\input{Peirce-popular-science-1.tex}
\input{Peirce-popular-science-2.tex}
\input{Peirce-popular-science-3.tex}

\end{document} 